---
title: "Table Parameters"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A11
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format:
  esqlabs-revealjs:
    scrollable: true
    code-overflow: scroll
---

::: {.r-fit-text}
## 🎯 Goal: Understand and Handle Table Parameters
:::

```{r setup, echo = FALSE}
#|output-location: default
knitr::opts_chunk$set(fig.showtext = TRUE, fig.align = "center")
```


# What are Table Parameters?

## Concept

- Table parameters in [OSPS](https://docs.open-systems-pharmacology.org/working-with-mobi/mobi-documentation/parameters-formulas-tags#table-formulas)

- Parameters can be defined as Tables made up out of pairs of x and y values.

- The parameter value will be linearly interpolated between the specified values. 

. . .

:::columns
:::{.column width="50%"}
Examples of parameters defined by a table:

- Enzyme ontogeny
- Compound solubility
:::

:::{.column width="50%"}
```{r, echo = FALSE}
ggplot2::ggplot(
  tibble::tibble(
    age = c(20, 30, 40, 50, 60, 70, 80),
    expression = c(0.0, 0.05, 0.10, 0.2, 0.4, 0.7, 1)
  ),
  ggplot2::aes(x = age, y = expression)
) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::labs(
    title = "Enzyme Ontogeny",
    x = "Age (years)",
    y = "Relative Expression"
  ) +
  ggplot2::theme_minimal()
```
:::
:::

## Load an Example

```{r}
library(ospsuite)
# Load a simulation
sim <- loadSimulation(system.file(
  "extdata",
  "simple.pkml",
  package = "ospsuite"
))

# Get the parameter defined by a table.
tableParam <- getParameter("Organism|TableParameter", sim)
```

# Access Table Values

## Inspect the Parameter

```{r}
tableParam <- getParameter("Organism|TableParameter", sim)
print(tableParam)
```
. . . 

```{r}
tableParam$value
tableParam$isTable
```

## All points

Points are stored under the `allPoints` attribute of `formula` within a parameter.

```{r}
tableParam$formula$allPoints
```

. . .

:::{.callout-note}
`allPoints` is a list of `ValuePoint` objects. Each has `$x`, `$y`, and `$restartSolver` fields.

```{r}
#| output-location: default
pt <- tableParam$formula$allPoints[[1]]
list(x = pt$x, y = pt$y, restartSolver = pt$restartSolver)
```
:::

## Specific points

```{r}
tableParam$formula$valueAt(xValue = 10)
```
. . .

```{r}
tableParam$formula$valueAt(xValue = 30)
```

## Linear Interpolation

`valueAt()` linearly interpolates between defined points.

```{r}
tableParam$formula$valueAt(xValue = 20)
```

. . .

```{r}
tableParam$formula$valueAt(xValue = 35)
```

# Modify Table Values

## Override with a single value

```{r}
setParameterValues(tableParam, 2)
tableParam
```

. . .

```{r}
tableParam$isFixedValue
```

## Reset to Original Values

`reset()` restores the table formula and clears the fixed-value override.

```{r}
tableParam$reset()
tableParam
tableParam$isFixedValue
```
## Add a new value 

You can add one or multiple points to the table 

```{r}
tableParam$formula$addPoints(5, 1.5)
tableParam
```

## Resources

-   [`{ospsuite}` vignette on table Parameters](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/table-parameters.html)

# Exercise

## Modify and Query a Table Parameter

1. Get the `Organism|TableParameter` from the simulation.
2. Check the value at `xValue = 25` (interpolated).
3. Override the table to a value of your choice 
4. Reset 


```{r}
#| code-fold: true
#| code-summary: "Solution"
# 1)
tableParam <- getParameter("Organism|TableParameter", sim)

# 2)
tableParam$formula$valueAt(xValue = 25)

# 3)
setParameterValues(tableParam, 2)

# 4)
tableParam$reset()
tableParam
```
