---
title: "Creating Individuals"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A12
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format:
  esqlabs-revealjs:
    scrollable: true
    code-overflow: scroll
---

::: {.r-fit-text}
## **🎯 Goal: Generate virtual individuals from the PK-Sim physiology database and apply them to a simulation**
:::

```{r setup, echo = FALSE}
#| output-location: default
knitr::opts_chunk$set(fig.showtext = TRUE, fig.align = "center")
suppressPackageStartupMessages(library(ospsuite))
options(ospsuite.plots.watermarkEnabled = FALSE)
```

## Overview

1. Define individual characteristics
2. Generate parameter values with `createIndividual()`
3. Apply parameters to a simulation
4. Assign enzyme ontogenies

# Define Individual Characteristics

## `createIndividualCharacteristics()`

Describes the individual before parameter generation.

Required:

- `species` — `Species$Human`, `Species$Rat`, `Species$Dog`, ...
- `population` (humans only) — `HumanPopulation$...`
- `gender` (humans only) — `Gender$Male` / `Gender$Female`

Optional:

- `weight`, `weightUnit` (default `"kg"`)
- `height`, `heightUnit` (default `"m"`)
- `age`, `ageUnit` (default `"year(s)"`)
- `moleculeOntogenies`

## Example

```{r}
#| output-location: default
library(ospsuite)

individualCharacteristics <- createIndividualCharacteristics(
  species = Species$Human,
  population = HumanPopulation$Japanese_Population,
  gender = Gender$Female,
  weight = 75,
  height = 1.75,
  heightUnit = "m",
  age = 43
)

individualCharacteristics
```

## Supported Species and Populations

```{r}
#| output-location: default
names(Species)
names(HumanPopulation)
```

# Generate Parameter Values

## `createIndividual()`

PK-Sim draws parameter values matching the characteristics.

```{r}
#| output-location: default
individual <- createIndividual(
  individualCharacteristics = individualCharacteristics
)

names(individual)
```

## Result Structure

- `distributedParameters` — parameters drawn from physiology distributions (`$paths`, `$values`)
- `derivedParameters` — formula-derived parameters (do **not** set manually)

```{r}
#| output-location: default
head(individual$distributedParameters$paths, 3)
head(individual$distributedParameters$values, 3)
```

:::{.callout-warning}
Only apply `distributedParameters` to the simulation. Setting `derivedParameters` breaks formula dependencies.
:::

# Apply to Simulation

## `setParameterValuesByPath()`

```{r}
#| output-location: default
simFilePath <- system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
)
sim <- loadSimulation(simFilePath)

setParameterValuesByPath(
  parameterPaths = individual$distributedParameters$paths,
  values = individual$distributedParameters$values,
  simulation = sim
)
```

:::{.callout-important}
**Species must match the original `.pkml`.** Applying a human individual to a rat simulation runs without error but produces nonsense.
:::

# Enzyme Ontogenies

## What Is a Molecule Ontogeny?

Molecule ontogenies describe the development of protein expression during aging.

The expression database stores ontogenies as age-dependent values.

When defining protein ontogenies for `individualCharacteristics`, the ontogeny value is calculated from the generated individual's age and stored in the parameters list.

## `MoleculeOntogeny`

Assign standard ontogeny patterns to custom molecules.

```{r}
#| output-location: default
myProteinOntogeny <- MoleculeOntogeny$new(
  molecule = "MyProtein",
  ontogeny = StandardOntogeny$CYP3A4
)

myProteinOntogeny
```

## Available Ontogenies

```{r}
#| output-location: default
names(StandardOntogeny)
```

## Attach to Characteristics

```{r, eval = FALSE}
#| output-location: default
individualCharacteristics <- createIndividualCharacteristics(
  species = Species$Human,
  population = HumanPopulation$European_ICRP_2002,
  gender = Gender$Male,
  age = 30,
  moleculeOntogenies = list(myProteinOntogeny)
)
```

# Exercise

## Exercise Overview

- Run the standard Aciclovir simulation (baseline)
- Create an Asian female individual, 45 yr, 60 kg, 1.65 m
- Apply individual parameters to the simulation and run again
- Compare both profiles in one figure

## Baseline Simulation

```{r, eval = TRUE}
#| output-location: default
#| code-fold: true
#| code-summary: "Solution"
simFilePath <- system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
)
simBaseline <- loadSimulation(simFilePath)
resultsBaseline <- runSimulations(simBaseline)[[1]]
```

## Create and Apply the Individual

```{r, eval = TRUE}
#| output-location: default
#| code-fold: true
#| code-summary: "Solution"
asianFemale <- createIndividualCharacteristics(
  species = Species$Human,
  population = HumanPopulation$Asian_Tanaka_1996,
  gender = Gender$Female,
  weight = 60,
  height = 1.65,
  heightUnit = "m",
  age = 45
)

indiv <- createIndividual(individualCharacteristics = asianFemale)

simIndiv <- loadSimulation(simFilePath)
setParameterValuesByPath(
  parameterPaths = indiv$distributedParameters$paths,
  values = indiv$distributedParameters$values,
  simulation = simIndiv
)

resultsAsian <- runSimulations(simIndiv)[[1]]
```

## Compare Profiles

```{r, eval = TRUE}
#| output-location: default
#| code-fold: true
#| code-summary: "Solution"
outputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"

dc <- DataCombined$new()
dc$addSimulationResults(
  simulationResults = resultsBaseline,
  quantitiesOrPaths = outputPath,
  names = "Baseline"
)
dc$addSimulationResults(
  simulationResults = resultsAsian,
  quantitiesOrPaths = outputPath,
  names = "Asian Female 45y"
)

plotTimeProfile(plotData = dc)
```

## Resources

- [`{ospsuite}` vignette on creating individuals](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-individual.html)
- [PK-Sim documentation on creating individuals](https://docs.open-systems-pharmacology.org/working-with-pk-sim/pk-sim-documentation/pk-sim-creating-individuals)