---
title: "Population Simulations"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A13
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---

::: {.r-fit-text}
## **🎯 Goal: Generate a virtual population and run a simulation across all individuals**
:::

```{r setup, echo = FALSE}
knitr::opts_chunk$set(fig.showtext = TRUE, fig.align = "center")
options(ospsuite.plots.watermarkEnabled = FALSE)
```

## Overview

1. Define population characteristics
2. Generate the population with `createPopulation()`
3. Export/import populations via CSV
4. Run a population simulation
5. Retrieve and plot results

# Preparation

## Load Sample Simulation

```{r}
library(ospsuite)

simFilePath <- system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
)
sim <- loadSimulation(simFilePath)
```

# Define Population Characteristics

## `createPopulationCharacteristics()`

Describes the population before sampling.

Key arguments:

- `species`, `population`
- `numberOfIndividuals`
- `proportionOfFemales` (percent, 0–100)
- `weightMin` / `weightMax` (`weightUnit`, default `"kg"`)
- `heightMin` / `heightMax` (`heightUnit`, default `"m"`)
- `ageMin` / `ageMax` (`ageUnit`, default `"year(s)"`)
- `moleculeOntogenies`

## Example

```{r}
popCharacteristics <- createPopulationCharacteristics(
  species = Species$Human,
  population = HumanPopulation$Asian_Tanaka_1996,
  numberOfIndividuals = 50,
  proportionOfFemales = 50,
  weightMin = 30,
  weightMax = 98,
  weightUnit = "kg",
  ageMin = 0,
  ageMax = 80,
  ageUnit = "year(s)"
)

popCharacteristics
```
The population argument selects a built-in physiological database from PK-Sim. Each entry in HumanPopulation corresponds to a published reference dataset (e.g., Asian_Tanaka_1996 is based on Tanaka et al. 1996). Run `names(HumanPopulation)` to see all available options.
# Generate Population

## `createPopulation()`

Samples all individuals at once.

```{r}
result <- createPopulation(
  populationCharacteristics = popCharacteristics
)

myPopulation <- result$population
myPopulation
```

## Inspect as Data Frame

```{r}
popDf <- populationToDataFrame(myPopulation)
dim(popDf)
head(popDf[, 1:5])
```

Each row = one individual; columns are parameter paths.

# CSV Round-Trip

## Export

```{r, eval = FALSE}
populationToCSV(
  population = myPopulation,
  filePath = "myPopulation.csv"
)
```

## Import

```{r, eval = FALSE}
loadedPop <- loadPopulation(csvPopulationFile = "myPopulation.csv")
```

:::{.callout-tip}
Use CSV to share populations across machines or reuse them without regenerating via PK-Sim.
:::

# Run Population Simulation

## `runSimulations()` with `population`

```{r}
populationResults <- runSimulations(
  simulations = sim,
  population = myPopulation
)[[1]]

populationResults
```

## Tune Parallelism

```{r, eval = FALSE}
simRunOptions <- SimulationRunOptions$new()
simRunOptions$numberOfCores <- 3
simRunOptions$showProgress <- TRUE

populationResults <- runSimulations(
  simulations = sim,
  population = myPopulation,
  simulationRunOptions = simRunOptions
)[[1]]
```

Defaults: `numberOfCores = max(cores) - 1`, `showProgress = FALSE`.

# Retrieve Results

## `getOutputValues()`

```{r}
resultsPath <- populationResults$allQuantityPaths[[1]]

resultsData <- getOutputValues(
  populationResults,
  quantitiesOrPaths = resultsPath
)

str(resultsData, max.level = 1)
names(resultsData$data)
head(resultsData$data)
```

Columns: `IndividualId`, `Time`, plus one column per requested path.

## Subset Individuals

```{r}
twoIndiv <- getOutputValues(
  populationResults,
  quantitiesOrPaths = resultsPath,
  individualIds = c(1, 2)
)

unique(twoIndiv$data$IndividualId)
```

# Visualize

## Quick Plot

```{r}
#| fig-height: 4
dc <- DataCombined$new()
dc$addSimulationResults(
  simulationResults = populationResults,
  quantitiesOrPaths = resultsPath,
  names = "Population"
)

ospsuite.plots::setDefaults()
plotTimeProfile(dc)
```

# Exercise

## Exercise Overview

- Create a 20-individual European population, ages 25–65, weights 55–95 kg
- Run the Aciclovir simulation for that population
- Plot plasma profiles of individuals 1 and 2

## Build Population and Run

```{r, eval = TRUE}
#| code-fold: true
#| code-summary: "Solution"
popChar <- createPopulationCharacteristics(
  species = Species$Human,
  population = HumanPopulation$European_ICRP_2002,
  numberOfIndividuals = 20,
  proportionOfFemales = 50,
  weightMin = 55,
  weightMax = 95,
  ageMin = 25,
  ageMax = 65
)

pop <- createPopulation(populationCharacteristics = popChar)$population

sim <- loadSimulation(simFilePath)
popResults <- runSimulations(simulations = sim, population = pop)[[1]]
```

## Plot Two Individuals

```{r, eval = TRUE}
#| code-fold: true
#| code-summary: "Solution"
path <- popResults$allQuantityPaths[[1]]

dc <- DataCombined$new()
dc$addSimulationResults(
  simulationResults = popResults,
  quantitiesOrPaths = path,
  individualIds = c(1, 2),
  names = "European Population"
)

plotTimeProfile(dc)
```

## Resources

- [`{ospsuite}` vignette on creating and running populations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-run-population.html)
- [PK-Sim populations documentation](https://docs.open-systems-pharmacology.org/working-with-pk-sim/pk-sim-documentation/pk-sim-creating-populations)
