---
title: "`{ospsuite}` Workflow"
subtitle: "Introduction to `{ospsuite}`"
module : "Introduction to `{ospsuite}`"
ID: A02
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---


::: {.r-fit-text}
## **🎯 Goal: Get a general overview of the `{ospsuite}` workflow**
:::


```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  fig.showtext = TRUE,
  fig.align = "center",
  out.width = "61.8%"
)

library(ospsuite)
options(ospsuite.plots.watermarkEnabled = FALSE)
```

# Introduction

## `{ospsuite}`

An R package that supports A-Z PK-Sim and MoBi simulation workflow on **R**.

-   Load
-   Manipulate
-   Run
-   Visualize
-   Calculate PK-parameters
-   Perform (local) sensitivity analysis

# R6 Reference Semantics

## R6 Reference Semantics {.scrollable}

Most `{ospsuite}` objects (`Simulation`, `Parameter`, `Molecule`, `SimulationResults`, `DataCombined`, ...) are **R6**: reference semantics, not base-R copy-on-modify.

-   Base R (data frames, vectors, lists): **copy on modify**
-   R6 objects: **mutate in place**; `x <- y` creates an alias, not a copy

```{r}
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)

a <- getParameter("Organism|Liver|Interstitial|Volume", sim)
b <- a # NOT a copy — same object

setParameterValues(b, 1.2)
a$value # 1.2 — mutation of b visible on a
```

Need fresh state? **Reload** the `.pkml` — `{ospsuite}` R6 objects are not cloneable.

## Working With R6 Objects

Access fields/methods with `$`; construct with `ClassName$new()`:

```{r, eval = FALSE}
sim$solver
sim$outputSelections

dc <- DataCombined$new()
dc$addSimulationResults(
  simulationResults = simulationResults,
  quantitiesOrPaths = outputPath
)
```

Practical rules:

-   `setParameterValues()`, `addOutputs()`, `clearOutputs()`, `addOutputInterval()` **mutate** their argument — no reassignment
-   Passing an R6 to a function = passing a pointer; the function can change it
-   `reset()` on a parameter restores its pkml default without reloading
-   Fresh independent state: call `loadSimulation(path)` again

# Basic workflow

## Load

```{r}
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
print(sim)
```

## Manipulate {.scrollable}

```{r}
parameterPath <- "Organism|Liver|Interstitial|Volume"
param <- getParameter(parameterPath, sim)
print(param)
```

```{r}
setParameterValues(param, 0.5)
print(param)
```

# R6 Reference Semantics

## Why This Matters

Most `{ospsuite}` objects (`Simulation`, `Parameter`, `Molecule`, `SimulationResults`, `DataCombined`, ...) are **R6** objects.

R6 behaves differently from base R:

-   Base R data frames, vectors, lists: **copy on modify** (functional style).
-   R6 objects: **reference semantics** (like Python, Java).

Notice what happened on the previous slide: `setParameterValues(param, 0.5)` did not return a new parameter. We never reassigned `param`. Yet the next `print(param)` shows the new value. The object was **mutated in place**.

## Assignment Does Not Copy

```{r}
a <- getParameter("Organism|Liver|Interstitial|Volume", sim)
b <- a # NOT a copy — b and a point to the same object

setParameterValues(b, 1.2)

a$value # also 1.2 — mutating b mutates a
```

Same applies to `sim`: `sim2 <- sim` gives two handles to one simulation. To work on an independent copy, **reload** the `.pkml`:

```{r}
sim_independent <- loadSimulation(simFilePath)
# sim and sim_independent are now fully separate
```

## Fields, Methods, and `$new()`

Access state and behavior with `$`:

```{r}
sim$solver # field
sim$outputSelections # field
```

Construct R6 objects with `ClassName$new(...)`:

```{r, eval = FALSE}
dc <- DataCombined$new()
sa <- SensitivityAnalysis$new(
  simulation = sim,
  parameterPaths = "Aciclovir|Lipophilicity"
)
```

Some methods are called via `$`:

```{r, eval = FALSE}
dc$addSimulationResults(
  simulationResults = simulationResults,
  quantitiesOrPaths = outputPath
)
```

## Practical Rules

-   Functions like `setParameterValues()`, `addOutputs()`, `clearOutputs()`, `addOutputInterval()` **mutate** their argument. No reassignment needed.
-   Passing an R6 object to a function is like passing a pointer — the function can change it.
-   `x <- y` where `y` is R6: `x` is an alias, not a copy.
-   Need a fresh state? Call `loadSimulation(path)` again — R6 objects in `{ospsuite}` are not cloneable.
-   `reset()` on a parameter restores its pkml default without reloading the whole simulation.

# Basic workflow

## Run

```{r}
# Always returns a named list
simulationResults <- runSimulations(sim)[[1]]
print(simulationResults)
# Convert simulation results to a data frame
head(simulationResultsToDataFrame(simulationResults))
```

## Visualize

```{r, echo = FALSE}
outputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
obsData <- lapply(
  c(
    "ObsDataAciclovir_1.pkml",
    "ObsDataAciclovir_2.pkml",
    "ObsDataAciclovir_3.pkml"
  ),
  function(x) {
    loadDataSetFromPKML(system.file("extdata", x, package = "ospsuite"))
  }
)
# Name the elements of the list according to the names of the loaded `DataSet`.
names(obsData) <- lapply(obsData, function(x) x$name)

myDataCombined <- DataCombined$new()

myDataCombined$addSimulationResults(
  simulationResults = simulationResults,
  quantitiesOrPaths = outputPath,
  names = "Aciclovir Plasma"
)

myDataCombined$addDataSets(
  obsData$`Vergin 1995.Iv`
)
```

```{r, out.width="61.7%"}
# Detailed description in following slides
plotTimeProfile(plotData = myDataCombined)
```


## Calculate PK-parameters {.scrollable}

```{r}
pkAnalyses <- calculatePKAnalyses(results = simulationResults)
head(pkAnalysesToDataFrame(pkAnalyses))
```

## Perform sensitivity analysis {.scrollable}

```{r}
sa <- SensitivityAnalysis$new(
  simulation = sim,
  parameterPaths = c(
    "Aciclovir|Lipophilicity",
    "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose"
  )
)
saResult <- runSensitivityAnalysis(sa)
print(saResult)
```


## Resources

-   [`{ospsuite}` vignettes](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/)
