---
title: "Load Models"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A03
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---

::: {.r-fit-text}
## **🎯 Goal: Learn how to load `.pkml` model files in `{ospsuite}`**
:::

## Included Example

`{ospsuite}` comes with an example simulation of Aciclovir in a `.pkml` file. It can be accessed with:

```{r, eval = FALSE}
#| output-location: default
system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
```

# Load

## Load a Model

Use the `loadSimulation()` function to load a `.pkml` file.

```{r}
library(ospsuite)
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

sim <- loadSimulation(simFilePath)

print(sim)
```

# Inspect

## Attributes

`Simulation` attributes can be browsed using `$`.


```{r}
# Solver settings
sim$solver
```

. . . 

```{r}
# Output paths
sim$outputSelections
```


. . .

```{r}
sim$outputSchema
```

## Simulation Tree

A `SimulationTree` can be generated using the `getSimulationTree()` function.

It will return a list of all the parameters available in the simulation.


```{r}
tree <- getSimulationTree(sim)

tree$Aciclovir[1:5]
```

---

```{r}
tree$Organism[1:5]
```

# Export

## Save a Model

`Simulation` objects can be exported to a `.pkml` file using the `saveSimulation()` function.

This is especially useful after having modified the original simulation object.

```{r, eval = FALSE}
saveSimulation(sim, "Aciclovir_modified.pkml")
```

# Exercise

---

## Load a `.pkml` file

1. Load the example simulation object included in the `{ospsuite}` package.
  Remember, you get its path using `system.file("extdata", "Aciclovir.pkml", package = "ospsuite")`.

```{r, eval = TRUE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answer"
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

sim <- loadSimulation(simFilePath)
```

---

## Parameter tree

1. Create the parameter tree for the simulation object
2. Explore the parameter tree within `Applications` to reach the `Volume` parameter in the simulation tree

```{r, eval = TRUE, results = FALSE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answer"
tree <- getSimulationTree(sim)

tree$Events$`IV 250mg 10min`$Application_1$Volume
```

