---
title: "Parameterize Simulations"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A04
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---

::: r-fit-text
## **🎯 Goal: Learn how to parameterize a model to perform new simulations using `{ospsuite}`**
:::

# Simulation Parameterization

## What Can Be Changed

Once a `.pkml` model file is loaded in `{ospsuite}`, simulation parameters can be changed from R.

Simulation parameters are accessed and modified using their **paths**.

Other aspects of the simulation that can be changed include:

-   Output Paths
-   Simulation time
-   Simulation outputs
-   Solver settings

# Paths & Containers

## Paths

Paths are character strings that represent the hierarchical structure of the simulation.

Each **element** of a path is separated by `|`.

```{r}
#| output-location: default
# molecule
molecule_path <- "Organism|VenousBlood|Plasma|Aciclovir"

# parameter
liver_interstitial_space <- "Organism|Liver|Interstitial|Volume"
```

## Paths in the Simulation Object

All parameter paths are stored in the simulation object.

```{r}
library(ospsuite)
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

sim <- loadSimulation(simFilePath)

tree <- getSimulationTree(sim)

tree$Organism$PeripheralVenousBlood$Aciclovir$`Plasma (Peripheral Venous Blood)`
```

## Path Anatomy

Parameter paths are organized in a hierarchical tree structure.

-   Each layer of a parameter path is a **container**.
-   The last element of a parameter path is a **parameter**, a **molecule amount**, an **observer**, or a **compartment**.

`Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)`

## Containers

A `Simulation` object is itself a container. But sub-containers can be created as path shortcuts with `getContainer()`.

```{r}
tree$Organism$Liver$Interstitial
```

. . .

```{r}
liver <- getContainer("Organism|Liver", sim)

liver$path
```

## Nested Containers

```{r}
liver_interstitial <- getContainer("Interstitial", liver)

liver_interstitial$path
```

## Explore Paths & Containers

`{ospsuite}` has special functions to explore the containers and paths of a simulation.

### `getAll*PathsIn()`

Return all the containers or parameters paths in a container.

```{r}
# get all the containers available in "Liver"
getAllContainerPathsIn(liver)[1:5]
```

## Parameter Paths in a Container

```{r}
getAllParameterPathsIn(liver)[1:5]
```

## Match Paths by Pattern

### `getAll*Matching()`

Return all the containers or parameters paths that match a specific pattern.

```{r}
head(
  getAllContainersMatching("Organism|*|Interstitial", sim),
  3
)
```

## Match Parameters by Pattern

```{r}
head(
  getAllParametersMatching("Organism|**|Volume", sim),
  3
)
```

It works in tandem with a special path syntax including *wildcards* (`*`).

## Wildcards

`**` is a wildcard that matches any structure of path elements.

In the example below, it means that we are looking for any number of containers between `Organism` and `Interstitial`.

```{r}
# get all the containers that have an Interstitial space with any amount of layers in between
interstitial_containers <- getAllContainersMatching(
  "Organism|**|Interstitial",
  sim
)
length(interstitial_containers)
```

. . .

```{r,results='hold'}
interstitial_containers[[1]]
interstitial_containers[[22]]
interstitial_containers[[9]]
```

## Single-Element Wildcard

`*` is a wildcard that matches exactly one path element.

```{r}
# get all the containers that have an Interstitial with exactly one path element before "Interstitial"
interstitial_containers_1 <- getAllContainersMatching(
  "Organism|*|Interstitial",
  sim
)
interstitial_containers_1[[1]]
```

. . .

```{r}
# get all the containers that have an Interstitial with exactly two path elements before "Interstitial"
interstitial_containers_2 <- getAllContainersMatching(
  "Organism|*|*|Interstitial",
  sim
)
interstitial_containers_2[[1]]
```

. . .

```{r}
# get all the containers that have an Interstitial with exactly three path elements before "Interstitial"
interstitial_containers_3 <- getAllContainersMatching(
  "Organism|*|*|*|Interstitial",
  sim
)
interstitial_containers_3[[1]]
```

# Parameter Values

## Get Parameter Values

Parameter values can be read from the simulation with `getParameter()`.

```{r}
getParameter("Organism|Liver|Interstitial|Volume", sim)
```

. . .

It can also be assigned for future access.

```{r}
liver_int_volume <- getParameter("Organism|Liver|Interstitial|Volume", sim)
```

## Sub-container Shortcut

::: {.callout-tip title="Tip: Identical results with sub-containers"}
``` r
getParameter("Interstitial|Volume", liver)
getParameter("Volume", liver_interstitial)
```
:::

## Set Parameters Values

Once a `get...()` function is used, the resulting object can be used to **change** the value of the parameter or molecule initial values.

```{r}
ageParam <- getParameter("Organism|Age", sim)
ageParam
```

. . .

```{r}
setParameterValues(ageParam, 55, units = "year(s)")
ageParam
```

## Set Values by Path

`setParameterValuesByPath()` is a shortcut that combines `getParameter()` and `setParameterValues()` into a single call. It accepts either a single path or a vector of paths.

```{r}
setParameterValuesByPath(
  parameterPaths = c("Organism|Age", "Organism|Weight"),
  values = c(40, 75),
  simulation = sim
)
```

## Reset Parameter Values

Original values can be restored with `reset()`.

```{r}
ageParam$reset()
ageParam
```

# Simulation Outputs

## Change Simulation outputs

The simulations outputs are set based on what is defined in the imported `.pkml` file.

```{r}
#| output-location: default
sim$outputSelections
```

. . .

Other simulation outputs can be added with `addOutputs()`

```{r}
addOutputs(
  "Organism|PeripheralVenousBlood|Aciclovir|Whole Blood (Peripheral Venous Blood)",
  simulation = sim
)

sim$outputSelections
```

## Clear Simulation Outputs

All simulation outputs can be removed with `clearOutputs()`.

```{r}
clearOutputs(sim)
# The outputs are now empty, we then add the original one

addOutputs(
  "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)",
  simulation = sim
)

sim$outputSelections
```

# Simulation time

## Change Simulation time

Default simulation time is the one defined in the imported `.pkml` file.

```{r}
#| output-location: default
sim$outputSchema
```

## Add an Output Interval

Another `OutputSchema` can be set with `addOutputInterval()`.

```{r}
addOutputInterval(
  simulation = sim,
  startTime = 1440,
  endTime = 1500,
  resolution = 2,
  intervalName = "highResEndSim"
)
sim$outputSchema
```

## Clear and Replace Intervals

All output intervals can be removed with the `clearOutputIntervals()` function.

```{r}
clearOutputIntervals(simulation = sim)
sim$outputSchema

#we add a new global interval
addOutputInterval(
  simulation = sim,
  startTime = 0,
  endTime = 1440,
  resolution = 0.5,
  intervalName = "globalRes"
)
sim$outputSchema
```

# Solver Settings

## Inspect Solver Settings

Numerical solver tolerances and step-size limits are stored in `sim$solver`. Tighter tolerances improve accuracy at the cost of runtime; loosen them only when runtime is prohibitive and accuracy loss is acceptable.

```{r}
sim$solver
```

## Change Solver Settings

Solver fields are writable directly. Typical knobs: `absTol` (absolute tolerance), `relTol` (relative tolerance), `h0` (initial step), `hMin` / `hMax` (step bounds), `mxStep` (maximum internal steps).

```{r}
sim$solver$absTol <- 1e-10
sim$solver$relTol <- 1e-9
sim$solver
```

# Exercise

## Exercise Overview

Walk through a full iterative modeling loop:

-   **Load the simulation.** Reload the Aciclovir `.pkml` into a fresh `sim` object.
-   **Find the application protocol paths.** List every parameter path under `Events`.
-   **Find the Dose parameter paths.** Filter the `Events` paths down to the `Dose` parameters.
-   **Change the Dose.** Create `sim_100mg` and set the dose to 100 mg.
-   **Change the administration protocol.** Set the infusion time to 30 minutes and verify the new value.
-   **Change the simulation time.** Replace the output intervals with a finer-then-coarser resolution schedule.

## Load Simulation

Let's load again the simulation in a new and fresh object called "sim"

```{r, eval = FALSE}
#| output-location: default
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

sim <- loadSimulation(simFilePath)
```

## Find the Application protocol paths

Let's have a look at the application protocol parameters

1.  Find the parameter paths that start with `Events`

```{r, results = FALSE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answer"
getAllParametersMatching("Events|**|*", sim)

# ** for any number of structure layer
# * for any parameter path endpoint
```

## Find the Dose parameter paths

Now, we anticipate that we will have to change the dose in our simulation. So we will need to identify all the `Dose` parameters within `Events`.

1.  Find all paths that start with `Events` and end with any kind of `Dose`

::: callout-tip
To look for all the `Dose` word, look for `*Dose*` in the path so that it will be found even if the word is surrounded by other characters.
:::

```{r, results = FALSE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answer"
getAllParametersMatching("Events|**|*Dose*", sim)

# ** for any number of structure layer
# * for any parameter path endpoint
# *Dose* to match any string containing "Dose".
```

## Change the Dose
Now, we will create another simulation with a different dose. Create one simulation object, `sim_100mg`, in which you will set the dose to 100 mg.

::: callout-tip
Once you have your sim object, you can either:
- use `getParameter()` then `setParameterValues()`, or
- use `setParameterValuesByPath()` directly as a shortcut
:::

```{r, results = FALSE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answer"
sim_100mg <- loadSimulation(simFilePath)

# Option 1: getParameter + setParameterValues
dose_100mg <- getParameter(
  "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  sim_100mg
)
setParameterValues(dose_100mg, 100, units = "mg")

# Option 2: setParameterValuesByPath
setParameterValuesByPath(
  parameterPaths = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  values = 100,
  units = "mg",
  simulation = sim_100mg
)
```
## Change Administration protocol {.scrollable}

Now, we need to set the drug infusion time to 30 minutes for our simulation.

1.  Find the parameter path for the `Infusion time` parameter in the `Applications` section.
2.  Set the parameter to 30 minutes.
3.  Get the value of the parameter to check if it has been changed.

```{r, results = FALSE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answer"
# 1) Infusion Time change
InfusionTime <- getParameter(
  "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Infusion time",
  sim_100mg
)
# 2)
setParameterValues(InfusionTime, 30, units = "min")

# 3)
getParameter(
  "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Infusion time",
  sim_100mg
)$value
```

## Change simulation time

We anticipate that the parameters we changed will mainly impact the first minutes of the simulation. So we want to increase the resolution for the first part of the simulation.

1.  Clear all previously defined `outputIntervals`
2.  Add a new interval from 0 to 15 minutes with 2 points per minutes.
3.  Add a new interval from 15 to 120 minutes with 1 point each 2 minutes
4.  Extend the simulation from 120 to 1440 minutes with 1 point each 15 minutes

## Change simulation time — Answers 1 & 2

```{r, results = FALSE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answers 1 & 2"
# 1)
clearOutputIntervals(simulation = sim_100mg)

# 2)
addOutputInterval(
  simulation = sim_100mg,
  startTime = 0,
  endTime = 15,
  resolution = 2,
  intervalName = "2 points per minute"
)
```

## Change simulation time — Answers 3 & 4

```{r, results = FALSE}
#| output-location: default
#| code-fold: true
#| code-summary: "Answers 3 & 4"
# 3)
addOutputInterval(
  simulation = sim_100mg,
  startTime = 15,
  endTime = 120,
  resolution = 0.5,
  intervalName = "1 point each 2 minutes"
)
# 4)
addOutputInterval(
  simulation = sim_100mg,
  startTime = 120,
  endTime = 1440,
  resolution = 0.07,
  intervalName = "1 point each 15 minutes"
)
```
