---
title: "Run Simulations"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A05
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---


::: {.r-fit-text}
## **🎯 Goal: Learn how to run simulations using `{ospsuite}`**
:::


# Run Simulations

## `runSimulations()`

Use `runSimulations()` for every run — one simulation or many.

- Takes a **list** of `Simulation` objects, or a single `Simulation` object.
- Returns a **named list** of `SimulationResults` objects, keyed by simulation id.

```r
runSimulations(sim)         # single
runSimulations(list(sim1, sim2))  # multiple
```

::: {.callout-warning title="Deprecated"}
`runSimulation()` (singular) is deprecated. Do not use it in new code.
:::

## Run a Single Simulation

```{r}
library(ospsuite)

sim <- loadSimulation(system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
))

simulationResults <- runSimulations(sim)

# runSimulations() always returns a list — unwrap to get the SimulationResults
simulationResult <- simulationResults[[1]]
simulationResult
```

## Run Multiple Simulations

```{r}
sim1 <- loadSimulation(system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
))
sim2 <- loadSimulation(system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
))

simulationResults <- runSimulations(simulations = list(sim1, sim2))

simulationResults
```

## Result List Names = Simulation IDs

The names of the list returned by `runSimulations()` are the **simulation ids** (`sim$id`), not the R variable names. Use them to look up a specific result.

```{r}
names(simulationResults)
```

. . .

```{r}
# Each simulation carries its own id
c(sim1$id, sim2$id)
```

. . .

```{r}
# Access a result by id
simulationResults[[sim1$id]]
```


# Simulation Results

## Access simulation results values

A `SimulationResults` object does not expose values directly. Extract them with one of the helpers below.

First, pick a single `SimulationResults` out of the list returned by `runSimulations()`.

```{r}
simulationResult <- simulationResults[[1]]
```

## `getOutputValues()`

```{r}
outputValues <- getOutputValues(simulationResult)
names(outputValues)
```

## Output Values — `data`

Contains a basic `data.frame` with simulation results values.

```{r}
outputValues$data
```

## Output Values — `metaData`

Contains context for the simulation results values.

```{r}
outputValues$metaData
```

## Pick Specific Outputs

`getOutputValues()` returns all outputs by default. Pass `quantitiesOrPaths` to restrict to one or more paths.

```{r}
plasmaPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
getOutputValues(simulationResult, quantitiesOrPaths = plasmaPath)$data
```

## `simulationResultsToDataFrame()`

Gathers simulation results values and metadata into a single `data.frame`.

```{r}
head(
  simulationResultsToDataFrame(simulationResult)
)
```


# Run Options

## `SimulationRunOptions`

`runSimulations()` accepts a `simulationRunOptions` argument built with `SimulationRunOptions$new()`. Use it to tune parallelism, control negative value checks, and more.

```{r, eval = TRUE}
options <- SimulationRunOptions$new(
  numberOfCores = 4, # run sims in parallel
  showProgress = TRUE, # print progress bar
  checkForNegativeValues = TRUE
)

simulationResults <- runSimulations(
  simulations = list(sim1, sim2),
  simulationRunOptions = options
)
```

Parallel execution matters when the list holds many simulations or running population simulations — scaling is near-linear up to physical core count.

## Handling Failed Runs

If a simulation fails, its slot in the returned list is `NULL`. Always check before extracting values.

```{r, eval = TRUE}
results <- runSimulations(list(sim1, sim2))

# Keep only successful runs
successful <- Filter(Negate(is.null), results)

# Or inspect which failed
failed_ids <- names(results)[vapply(results, is.null, logical(1))]
```

# Save & Load Results

## Export to CSV

`exportResultsToCSV()` writes a `SimulationResults` object to disk. Useful for caching long-running simulations or sharing with colleagues.

```{r, eval = FALSE}
exportResultsToCSV(simulationResult, "aciclovir_results.csv")
```

## Import from CSV

`importResultsFromCSV()` reconstructs a `SimulationResults` object from a previously exported CSV. Requires the original `Simulation` object.

```{r, eval = FALSE}
simulationResult <- importResultsFromCSV(sim, "aciclovir_results.csv")
```


# Exercise

## Exercise Overview

Compare two doses of Aciclovir end-to-end:

-   **Load & parameterize.** Two simulations, one with the default dose and one at 100 mg.
-   **Run both.** A single call to `runSimulations()`.
-   **Extract results.** Pull the output values out of the returned list.
-   **Compare.** Maximum plasma concentration for each run.

## Load and create two simulations {.scrollable}

1. Load the example Aciclovir model to create two simulations: `sim` and `sim_100mg`
2. Set `Dose` parameter of `sim_100mg` to 100 mg (path: `Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose`)
3. Run both simulations


```{r, eval = TRUE}
#| code-fold: true
#| code-summary: "Answer 1 & 2"
library(ospsuite)

# 1)
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

sim <- loadSimulation(simFilePath)
sim_100mg <- loadSimulation(simFilePath)

# 2)
setParameterValuesByPath(
  parameterPaths = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  values = 100,
  units = "mg",
  simulation = sim_100mg
)
```

## Run Both Simulations

```{r, eval = TRUE}
#| code-fold: true
#| code-summary: "Answer 3"
# 3)
simulationResults <- runSimulations(list(sim, sim_100mg))
```

## Extract the simulation results {.scrollable}

1. Extract the output values of `sim` and `sim_100mg` and assign them to `results_sim` and `results_sim_100mg` respectively.
2. Get the maximum concentration of Aciclovir in the Plasma for both simulations.

```{r, eval = TRUE, results = FALSE}
#| code-fold: true
#| code-summary: "Answer"
# 1)
names(simulationResults)

results_sim <- simulationResultsToDataFrame(simulationResults[[sim$id]])
results_sim_100mg <- simulationResultsToDataFrame(simulationResults[[
  sim_100mg$id
]])

# 2) Aciclovir plasma is the only output in the default pkml,
#    so max() over simulationValues is the plasma Cmax.
max(results_sim$simulationValues)
max(results_sim_100mg$simulationValues)

#> 50.25272
#> 20.10109
```
