---
title: "Plot Simulation Results"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A06
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---




::: {.r-fit-text}
## **🎯 Goal: Learn how to create plots from Simulation Results using `{ospsuite}`**
:::


```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  fig.showtext = TRUE,
  fig.align = "center",
  out.width = "61.8%"
)

library(ospsuite)
ospsuite.plots::setDefaults()
options(ospsuite.plots.watermarkEnabled = FALSE)
```

# Plot Simulation Results

## Simulated and Observed Data

Comparing simulated and observed data is crucial in modeling and simulation workflow.

`{ospsuite}` provides a set of features to handle these data before visualizing them.

-   `DataSet` to import and store observed data,
-   `DataCombined` to link observed and simulated data

## DataSet

In `{ospsuite}` observed data are stored in `DataSet` objects.

Standard format for data coming from:

-   Manual entry,
-   `.pkml` files,
-   Excel files

## Create DataSet in R

Each `DataSet` must be **named** when the object is created.

```{r}
dataSet <- DataSet$new("Aciclovir")
dataSet
```

## Set Values

```{r}
dataSet$setValues(
  xValues = c(1, 2, 3, 4),
  yValues = c(0, 0.1, 0.6, 10),
  yErrorValues = c(0.001, 0.001, 0.1, 1)
)
dataSetToTibble(dataSet)
```

## Import DataSet from PKML Files

```{r}
filePath <- system.file(
  "extdata",
  "ObsDataAciclovir_1.pkml",
  package = "ospsuite"
)
dataSet <- loadDataSetFromPKML(filePath = filePath)
dataSetToTibble(dataSet)
```

## Import DataSet from Excel File

Loading observed data from an Excel sheet requires an `ImporterConfiguration` object.

`ImporterConfiguration` describes the mapping of the columns and metadata of the Excel sheet.

Can be created in PK-Sim or MoBi (`.xml` file) and loaded in R using the `loadDataImporterConfiguration()` function.

## Load Importer Configuration {.scrollable}

```{r}
configurationFilePath <- system.file(
  "extdata",
  "dataImporterConfiguration.xml",
  package = "ospsuite"
)
excelFilePath <- system.file(
  "extdata",
  "CompiledDataSet.xlsx",
  package = "ospsuite"
)
dataImporterConfiguration <- loadDataImporterConfiguration(
  configurationFilePath = configurationFilePath
)

dataImporterConfiguration
```

## Load Data Sets from Excel

```{r}
dataImporterConfiguration$sheets <- "TestSheet_1"
dataSets <- loadDataSetsFromExcel(
  xlsFilePath = excelFilePath,
  importerConfigurationOrPath = dataImporterConfiguration
)

names(dataSets)
```

## DataCombined

Links simulated and observed data.

✅ Simulated Data + ✅ Observed Data = `DataCombined`

## Create a DataCombined Object

```{r}
myDataCombined <- DataCombined$new()
myDataCombined
```


## Add simulation results

Use the `addSimulationResults()` method.

```{r}
sim <- loadSimulation(system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
))

clearOutputs(sim)
addOutputs(
  "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)",
  sim
)

sim_res <- runSimulations(sim)

myDataCombined$addSimulationResults(
  simulationResults = sim_res[[1]]
)
```

. . .

## Add simulation results — Inspect

```{r}
myDataCombined
```

## Add Observed Data

Use the `addDataSets()` method.

```{r}
filePath <- system.file(
  "extdata",
  "ObsDataAciclovir_1.pkml",
  package = "ospsuite"
)
obsData <- loadDataSetFromPKML(filePath = filePath)

myDataCombined$addDataSets(
  obsData
)
myDataCombined
```

## Inspect as Data Frame

```{r}
myDC_df <- myDataCombined$toDataFrame()
head(myDC_df)
```

. . .

```{r}
unique(myDC_df$name)
```

## First Look — Default Plot

```{r}
plotTimeProfile(myDataCombined)
```

## Rename and Group Simulated and Observed Data

```{r}
myDataCombined <- DataCombined$new()

myDataCombined$addSimulationResults(
  simulationResults = sim_res[[1]],
  names = "Simulated",
  groups = "Aciclovir"
)

myDataCombined$addDataSets(
  obsData,
  names = "Observed (Vergin 1995.Iv)",
  groups = "Aciclovir"
)

myDC_df <- myDataCombined$toDataFrame()
```

## Rename and Group — Inspect

```{r}
unique(myDC_df$name)
```

. . .

```{r}
unique(myDC_df$group)
```

. . .

```{r, echo = TRUE}
plotTimeProfile(myDataCombined)
```

## Add several Observed Data

```{r, warning= FALSE}
myDataCombined2 <- DataCombined$new()


myDataCombined2$addSimulationResults(
  simulationResults = sim_res[[1]],
  names = "Simulated"
)

obsData <- lapply(
  c(
    "ObsDataAciclovir_1.pkml",
    "ObsDataAciclovir_2.pkml",
    "ObsDataAciclovir_3.pkml"
  ),
  function(x) {
    loadDataSetFromPKML(system.file("extdata", x, package = "ospsuite"))
  }
)
# Name the elements of the list according to the names of the loaded `DataSet`.
names(obsData) <- lapply(obsData, function(x) x$name)

myDataCombined2$addDataSets(
  dataSets = c(obsData$`Vergin 1995.Iv`, obsData$`Laskin 1982.Group C`),
  names = c("Observed (Vergin 1995.Iv)", "Observed (Laskin 1982.Group C)"),
  groups = "Aciclovir"
)

myDC_df <- myDataCombined2$toDataFrame()
```

## Several Observed — Inspect

```{r, warning= FALSE}
unique(myDC_df$name)
```

. . .

```{r, warning= FALSE}
unique(myDC_df$group)
```

## Several Observed — Plot

```{r}
plotTimeProfile(myDataCombined2)
```

## Generate Plots

`{ospsuite.plots}` provides five plotting functions. All return `ggplot2` objects.

-   `plotTimeProfile()` — time profiles
-   `plotPredictedVsObserved()` — goodness of fit
-   `plotResidualsVsCovariate()` — residuals against time, observed, or predicted
-   `plotResidualsAsHistogram()` — residuals distribution
-   `plotQuantileQuantilePlot()` — residuals normality check

::: {.callout-note title="`setDefaults()`"}
`ospsuite.plots::setDefaults()` (called in the setup chunk) installs the OSP house style: color palettes, linear/log scales, legend placement, and watermark settings. Call it once per session before plotting.
:::

## Time Profile

```{r}
plotTimeProfile(myDataCombined)
```

## Time Profile — Log Scale

Plasma profiles are typically read on a log scale. Pass `yScale = "log"`.

```{r}
plotTimeProfile(myDataCombined, yScale = "log")
```

## Predicted vs Observed

```{r}
plotPredictedVsObserved(myDataCombined)
```

## Residuals vs Predicted

```{r}
plotResidualsVsCovariate(myDataCombined, xAxis = "predicted")
```

## Residuals vs Time

```{r}
plotResidualsVsCovariate(myDataCombined, xAxis = "time")
```

## Residuals Histogram

```{r}
plotResidualsAsHistogram(myDataCombined)
```

## Residuals Q-Q Plot

```{r}
plotQuantileQuantilePlot(myDataCombined)
```

## Customize Plots

All plotting functions return `ggplot2` objects, so customization uses the standard `ggplot2` `+` operator.
`ggplot2` builds plots in layers: each `+` adds something new (a title, a theme, axis limits, etc.). A quick reference: [ggplot2.tidyverse.org](https://ggplot2.tidyverse.org/reference/index.html)

```{r}
library(ggplot2)

plotTimeProfile(myDataCombined) +
  labs(
    title = "Example: Customize a Plot",
    subtitle = "Using ggplot2 layers",
    caption = "Source: Aciclovir data"
  ) +
  theme_minimal()
```

## Multi-Panel with `{patchwork}`

`{ospsuite.plots}` returns plain `ggplot2` objects. Combine them with `{patchwork}` operators `+`, `|`, `/`;
`|` places plots side by side, `/` stacks them vertically.

```{r}
library(patchwork)

timeProfile <- plotTimeProfile(myDataCombined)
resVsPred <- plotResidualsVsCovariate(myDataCombined, xAxis = "predicted")
resVsTime <- plotResidualsVsCovariate(myDataCombined, xAxis = "time")

((timeProfile | resVsPred) / resVsTime) +
  plot_annotation(title = "Multiple plots in one figure")
```

## Save Plots

Use `ospsuite.plots::exportPlot()` to write a plot to disk.

```{r, eval = FALSE}
p <- plotTimeProfile(myDataCombined)

ospsuite.plots::exportPlot(
  plotObject = p,
  filePath = "resources/A6/exportedPlot.png",
  width = 20,
  height = NULL,
  dpi = 300
)
```


# Exercises

## Run Simulations

1. Load ospsuite and the aciclovir model in two separate simulation objects called `sim` and `sim_100mg`
2. Modify the `Dose` parameter in `sim_100mg` to 100 mg
3. Run the simulations and assign the output to `results`

```{r, eval=TRUE}
#| code-fold: true
#| code-summary: "Answer"
library(ospsuite)
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

sim <- loadSimulation(simFilePath)
sim_100mg <- loadSimulation(simFilePath)

setParameterValuesByPath(
  simulation = sim_100mg,
  parameterPaths = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  values = 100,
  units = "mg"
)

results <- runSimulations(simulations = list(sim = sim, sim_100mg = sim_100mg))
```


## Create a `DataCombined` object

1. Initialize a `DataCombined` object called `myDataCombined`,
2. Load the data set contained in the `ObsDataAciclovir_1.pkml` file,
3. Add the dataset to `myDataCombined` and specify that the data set name is `Observed` and that it belongs to the `Aciclovir` group,


```{r, eval=TRUE}
#| code-fold: true
#| code-summary: "Answer"
# 1)
myDataCombined <- DataCombined$new()

# 2)
filePath <- system.file(
  "extdata",
  "ObsDataAciclovir_1.pkml",
  package = "ospsuite"
)
dataSet <- loadDataSetFromPKML(filePath = filePath)

# 3)

myDataCombined$addDataSets(
  dataSet,
  names = "Observed",
  groups = "Aciclovir"
)
```


## Add Simulation Results

1. Add the simulation results to `myDataCombined` and specify that the simulation results name is `Simulated` and that it belongs to the `Aciclovir` group.
2. Plot the time profile for `myDataCombined`.

::: {.callout-tip}
The default pkml exposes more than one output path. When passing a single `name`, also pass `quantitiesOrPaths` to restrict to one output — the vector lengths must match.
:::

```{r, eval=TRUE}
#| code-fold: true
#| code-summary: "Answer"
plasmaPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"

myDataCombined$addSimulationResults(
  simulationResults = results[[1]],
  quantitiesOrPaths = plasmaPath,
  names = "Simulated",
  groups = "Aciclovir"
)

plotTimeProfile(myDataCombined)
```
## Update Simulation Results

1. Replace previous simulation results with the results of the `sim_100mg` simulation in `myDataCombined`
2. Plot the time profile and the predicted vs observed plot from `myDataCombined`

```{r, eval=TRUE}
#| code-fold: true
#| code-summary: "Answer"
myDataCombined$addSimulationResults(
  simulationResults = results[[2]],
  quantitiesOrPaths = plasmaPath,
  names = "Simulated",
  groups = "Aciclovir"
)


plotTimeProfile(myDataCombined)
plotPredictedVsObserved(myDataCombined)
```

## Plot several outputs

1. Clear the default outputs of `sim` and set the following three outputs:
    - `Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)`
    - `Organism|PeripheralVenousBlood|Aciclovir|Whole Blood (Peripheral Venous Blood)`
    - `Organism|Kidney|Plasma|Aciclovir|Concentration`

2. Rerun the simulation
3. Create a new `DataCombined` object called `myDataCombined2`

## Plot several outputs — Answer

```{r, eval=TRUE}
#| code-fold: true
#| code-summary: "Answer"
# 1) clear default outputs first, then add the three we want
clearOutputs(sim)
addOutputs(
  quantitiesOrPaths = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)",
  simulation = sim
)
addOutputs(
  quantitiesOrPaths = "Organism|PeripheralVenousBlood|Aciclovir|Whole Blood (Peripheral Venous Blood)",
  simulation = sim
)
addOutputs(
  quantitiesOrPaths = "Organism|Kidney|Plasma|Aciclovir|Concentration",
  simulation = sim
)

# 2)
results2 <- runSimulations(sim)

# 3)
myDataCombined2 <- DataCombined$new()
```


## Add Simulation Results and Customize plot

4. Add the simulation results to `myDataCombined2`. The number of names to provide (in a vector) must match the number of `results2[[1]]$allQuantityPaths` elements.
5. Plot the time profile for `myDataCombined2`, restrict the x-axis to 0–60 minutes, and add a title — using `ggplot2` layers on the returned plot.

## Add Simulation Results and Customize plot — Answer

```{r, eval=TRUE}
#| code-fold: true
#| code-summary: "Answer"
library(ggplot2)

# 4)
myDataCombined2$addSimulationResults(
  simulationResults = results2[[1]],
  names = c("Plasma", "Whole Blood", "Kidney Plasma")
)

# 5)
plotTimeProfile(myDataCombined2) +
  coord_cartesian(xlim = c(0, 60)) +
  labs(
    title = "Comparison of Aciclovir Concentrations in Different Compartments"
  )
```

