---
title: "Dimensions & Units"
subtitle: "Introduction to `{ospsuite}`"
module: "Introduction to `{ospsuite}`"
ID: A07
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---

::: {.r-fit-text}
## **đźŽŻ Goal: Learn how to wrangle with units**
:::

```{r setup, echo = FALSE}
knitr::opts_chunk$set(fig.showtext = TRUE, fig.align = "center")
```

# Dimensions and Units

## In PK-Sim and MoBi

PK-Sim uses the concept of Dimensions and Units

- A **dimension** represents the type of quantity (*e.g.* Volume, Mass, Concentration)
- A **unit** is one of several ways to measure a dimension (*e.g.* L, mL; kg, g; mg/L, Âµg/mL)

. . .

The software uses a **base unit** for each dimension

- Volume is in L
- Mass is in kg
- Concentration (molar) is in Âµmol/L

## Base vs Display Units

:::{.callout-note}
Although users can set or visualize a quantity in different units, `OSPS` always uses the base unit for calculations.
:::

. . .

:::{.callout-important}
The base units are often different from the units displayed by default in PK-Sim. Those are called the **display units**.
:::

. . .

[Full List of Dimensions and Units](https://docs.open-systems-pharmacology.org/appendix/appendix#a.1.-dimensions-and-base-units)

## In `{ospsuite}`

`{ospsuite}` uses the same concept of Dimensions and Units.

All Dimensions and Units available are listed in the package.


```{r, echo = FALSE}
library(ospsuite)
```

```{r, out.height="20%"}
ospsuite::ospDimensions
```

## Available Units

```{r, out.height="20%"}
ospsuite::ospUnits
```


# Base and Display Units

## Base Units Internally

Like in PK-Sim and MoBi, `{ospsuite}` uses base units internally.
Users interact with the package through display units or any other supported unit.

```{r}
#| output-location: fragment
library(ospsuite)

sim <- loadSimulation(system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
))

bmiParam <- getParameter("Organism|BMI", sim)

bmiParam
```

## Unit Fields on a Parameter

:::{.callout-note title="Base and Display units are listed in Parameters"}
```{r}
bmiParam$unit
bmiParam$displayUnit
bmiParam$allUnits
```
:::

kg/dmÂ˛ is the base unit for BMI. Display unit is kg/mÂ˛.

# Convert Units

## `{ospsuite}` Unit Conversion Functions

The package provides a set of functions for easy conversion between units of a same dimension.

. . .

- `toUnit()` to convert from one unit to another
- `toBaseUnit()` to convert from any unit to the base unit
- `toDisplayUnit()` to convert from the base unit to the display unit

. . .

```{r}
ospsuite::toUnit(
  quantityOrDimension = "Volume",
  values = 1,
  sourceUnit = "L",
  targetUnit = "mL"
)
```

. . .

```{r}
ospsuite::toBaseUnit(quantityOrDimension = "Volume", values = 1000, unit = "mL")
```

## Convert to Display Unit

```{r}
bmiParam
```

. . .

```{r}
toDisplayUnit(quantity = bmiParam, values = 0.23)
```

## Additional Arguments

In some situations, the `molWeight` and `molWeightUnit` arguments are required to convert between Weight and Molar Concentrations.

. . .

```{r}
ospsuite::toUnit(
  quantityOrDimension = "Concentration (mass)",
  values = 100,
  sourceUnit = "mg/L",
  targetUnit = "mol/L",
  molWeight = 225.21,
  molWeightUnit = "g/mol"
)
```


## Set Parameter Values

The function `setParameterValues()` can be used to set the value of a parameter from a specific unit.

```{r}
setParameterValues(bmiParam, 25)

bmiParam
```

. . .

```{r}
setParameterValues(bmiParam, 25, units = "kg/mÂ˛")
bmiParam
```

## Convert Then Set

```{r}
setParameterValues(bmiParam, toBaseUnit(bmiParam, 25, unit = "kg/mÂ˛"))
bmiParam
```


## DataCombined

A `DataCombined` may hold simulation results and observed data in different units. Use `convertUnits()` to align them, or to change units for reporting.

```{r}
sim_res <- runSimulations(sim)[[1]]
filePath <- system.file(
  "extdata",
  "ObsDataAciclovir_1.pkml",
  package = "ospsuite"
)
dataSet <- loadDataSetFromPKML(filePath = filePath)

myDataCombined <- DataCombined$new()
myDataCombined$addSimulationResults(
  simulationResults = sim_res
)
myDataCombined$addDataSets(
  dataSet,
  names = "Observed (Vergin 1995.Iv)",
  groups = "Aciclovir"
)
```

## Inspect Original Units

```{r}
myDataCombined$toDataFrame()
```

Original data has Time in minutes and Concentration in Âµmol/mL.

## `convertUnits()`

The function `convertUnits()` can be used to convert the units of the DataCombined.


```{r}
convertUnits(myDataCombined, xUnit = "h", yUnit = "mg/L")
```

# Exercises

## Exercise Overview

Practice unit handling on the Aciclovir Dose parameter:

- **Access the Dose parameter.** Use `getParameter()`.
- **Inspect units.** Base unit and display unit.
- **Convert kg â†’ mg.** Try `toUnit()` and `toDisplayUnit()`.
- **Change value.** Set Dose to 100 mg with `setParameterValues()`.
- **Verify.** Check if the stored unit changed.

## Access the Dose Parameter

Access the administration Dose from the simulation. What is the base unit for this parameter?

```{r}
sim <- loadSimulation(system.file(
  "extdata",
  "Aciclovir.pkml",
  package = "ospsuite"
))
```

```{r}
#| code-fold: true
#| code-summary: "Solution"
Dose <- getParameter(
  "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  sim
)

Dose$unit
```

. . .

What is the display unit of the Dose?

```{r}
#| code-fold: true
#| code-summary: "Solution"
Dose$displayUnit
```


:::notes
The base unit is kg and the display unit is mg.
:::

## Convert kg to mg

By using two functions of your choice among `toUnit()`, `toBaseUnit()` and `toDisplayUnit()`, convert the Dose value from kg to mg.

```{r}
#| code-fold: true
#| code-summary: "Solution"
# We can use toUnit and toDisplayUnit
ospsuite::toUnit(
  quantityOrDimension = Dose,
  values = Dose$value,
  sourceUnit = "kg",
  targetUnit = "mg"
)
```


. . .

```{r}
#| code-fold: true
#| code-summary: "Solution"
ospsuite::toDisplayUnit(quantity = Dose, values = Dose$value)
```

. . .

Then, use `setParameterValues()` to change the value to 100 mg.

```{r}
#| code-fold: true
#| code-summary: "Solution"
setParameterValues(Dose, 100, units = "mg")
```

## Verify Stored Unit

Display the Dose object. Has the unit changed?

```{r}
#| code-fold: true
#| code-summary: "Solution"
Dose
```


:::notes
The unit is still in kg, as the base unit remains the same.
:::



# Resources

## Further Reading

- [Full list of Dimensions and Base Units](https://docs.open-systems-pharmacology.org/appendix/appendix#a.1.-dimensions-and-base-units)
- [`{ospsuite}` vignettes](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/unit-conversion.html)
