---
title: "Project Initialization"
subtitle: "Introduction to `{esqlabsR}`"
module: "Introduction to `{esqlabsR}`"
ID: B2
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format:
  esqlabs-revealjs:
    scrollable: true
    code-overflow: scroll
---

::: {.r-fit-text}
## 🎯 Goal: Initialize a {esqlabsR} project
:::

# Introduction

```{r, include = FALSE}
library(esqlabsR)
```

## Project Structure

::::: columns
::: {.column width="40%"}
```
 Project Folder
 ├── ProjectConfiguration.xlsx
 ├── ProjectConfiguration.json
 ├── Data/
 ├── Models/
 ├── Configurations/
 └── Results/
```
:::

::: {.column width="60%"}
- 4 folders.
- 1 `ProjectConfiguration.xlsx` to link them all.
- 1 `ProjectConfiguration.json` snapshot of the project state.
:::
:::::

## `ProjectConfiguration.xlsx`

Tells `{esqlabsR}` where to find each folder/file. Ships with sensible defaults; customize to match your project layout.
This makes the workflow fully reproducible: anyone with the same files and configuration will get the same results.

Load the bundled example to inspect:

```{r}
my_project_configuration <- createProjectConfiguration(
  path = esqlabsR::exampleProjectConfigurationPath(),
  ignoreVersionCheck = TRUE
)
my_project_configuration
```

## Properties — Folders

| Property | Description |
|---|---|
| `projectConfigurationFilePath` | Path to your `ProjectConfiguration.xlsx`. Anchors all relative paths. |
| `projectConfigurationDirPath` | Folder where that file lives. |
| `modelFolder` | Folder with `.pkml` simulation files. |
| `configurationsFolder` | Folder with the configuration Excel files. |
| `dataFolder` | Folder with experimental data files. |
| `outputFolder` | Folder where simulation results are written. |

## Properties — Configuration Files

| Property | Description |
|---|---|
| `modelParamsFile` | Excel file with model parameter overrides. |
| `individualsFile` | Excel file with individual characteristics. |
| `populationsFile` | Excel file describing populations. |
| `populationsFolder` | Folder with populations defined as CSV files. |
| `scenariosFile` | Excel file with scenario definitions. |
| `applicationsFile` | Excel file with application protocols. |
| `plotsFile` | Excel file with plot definitions. |
| `parameterIdentificationFile` | Excel file with parameter-identification tasks. |

## Properties — Data & Status

| Property | Description |
|---|---|
| `dataFile` | Excel file with experimental data. |
| `dataImporterConfigurationFile` | XML config used to import the data. |
| `modified` | Whether any property has changed since loading. Read-only. |
| `esqlabsRVersion` | Version of `{esqlabsR}` recorded in the project. Read-only. |

# Project Folders

## Data Folder

```{.bash code-line-numbers="4"}
 Project Folder
 ├── ProjectConfiguration.xlsx
 ├── ProjectConfiguration.json
 ├── Data/
 ├── Models/
 ├── Configurations/
 └── Results/
```

::: {.center-x}
Where to store all input data for the project.
:::

## Data Templates

The **Data** folder also contains some input templates:

:::::: columns
::: {.column width="40%"}
```
Project Folder/Data/
├── *_TimeValuesData.xlsx
└── dataImporter_configuration.xml
```
:::

:::: {.column width="60%"}
::: incremental
- `*_TimeValuesData.xlsx` — sample file with the expected structure for observed data.
- `dataImporter_configuration.xml` — sample import configuration matching the `*_TimeValuesData.xlsx` structure.
- Can also be loaded in PK-Sim and MoBi.
:::
::::
::::::

## Models Folder

```{.bash code-line-numbers="5"}
 Project Folder
 ├── ProjectConfiguration.xlsx
 ├── ProjectConfiguration.json
 ├── Data/
 ├── Models/
 ├── Configurations/
 └── Results/
```

Stores all simulation files coming from PK-Sim or MoBi.

:::::: columns
::: {.column width="40%"}
```
Project Folder/Models/
└── Simulations/
```
:::

::: {.column width="60%"}
- `Simulations/` — where to store `.pkml` simulation files.
:::
::::::

## Configurations Folder

```{.bash code-line-numbers="6"}
 Project Folder
 ├── ProjectConfiguration.xlsx
 ├── ProjectConfiguration.json
 ├── Data/
 ├── Models/
 ├── Configurations/
 └── Results/
```

Contains all the essential settings files for the `{esqlabsR}` workflow.

## Configuration Files

:::::: columns
::: {.column width="40%"}
```
Project Folder/Configurations/
├── Applications.xlsx
├── Individuals.xlsx
├── ModelParameters.xlsx
├── ParameterIdentification.xlsx
├── Plots.xlsx
├── Populations.xlsx
├── Scenarios.xlsx
└── PopulationsCSV/
    └── TestPopulation.csv
```
:::

:::: {.column width="60%"}
::: incremental
- `Applications.xlsx` — define administration protocols.
- `Individuals.xlsx` — set up individual characteristics.
- `ModelParameters.xlsx` — specify model parameter sets.
- `Populations.xlsx` + `PopulationsCSV/` — define populations.
- `Scenarios.xlsx` — main file defining simulation **scenarios** using all the previously listed files.
- `Plots.xlsx` — define plots and plot grids that will be generated.
- `ParameterIdentification.xlsx` — define PI tasks.
:::
::::
::::::

## Results Folder

```{.bash code-line-numbers="7"}
 Project Folder
 ├── ProjectConfiguration.xlsx
 ├── ProjectConfiguration.json
 ├── Data/
 ├── Models/
 ├── Configurations/
 └── Results/
```

Default destination for `SimulationResults` and generated plots.

# `ProjectConfiguration.json`

## Snapshot of the Project

::: {.center-x}
- Captures `ProjectConfiguration.xlsx` and all Excel files in `Configurations/`.
- Plain text — version-control friendly.
- Can be committed to Git and shared with collaborators.

Same purpose as `ProjectConfiguration.xlsx`, but in plain text: easier to track changes, compare versions, and collaborate via Git.
:::

## Validate

`validateAllConfigurations()` checks the project for structural issues, data integrity, uniqueness, and cross-file references.

```r
validationResults <- validateAllConfigurations(myProjectConfig)

if (isAnyCriticalErrors(validationResults)) {
  validationSummary(validationResults)
}
```

::: {.callout-note}
Each entry of `validationResults` exposes `critical_errors`, `warnings`, and `data` — use `validationSummary()` for a readable overview.
:::

## Excel ↔ JSON Interoperability

| Function | Direction | Purpose |
|---|---|---|
| `snapshotProjectConfiguration()` | Excel → JSON | Freeze current state into a snapshot. |
| `restoreProjectConfiguration()` | JSON → Excel + folders | Recreate Excel files (and folder layout) from a snapshot. |
| `projectConfigurationStatus()` | compare | Report whether Excel and JSON are in sync. |

## Round-Trip Example

```r
# 1. Snapshot the current Excel state
snapshotProjectConfiguration("ProjectConfiguration.xlsx")

# 2. Commit ProjectConfiguration.json to Git, share with team

# 3. Restore Excel files on another machine
restoredConfig <- restoreProjectConfiguration("ProjectConfiguration.json")

# 4. Check sync status at any time
status <- projectConfigurationStatus(
  "ProjectConfiguration.xlsx",
  "ProjectConfiguration.json"
)
```

# Initialize a Project

## `initProject()`

`{esqlabsR}` ships a default project structure. Scaffold it with one call:

```{r, eval = FALSE}
#| output-location: default
esqlabsR::initProject(destination = "path/to/your/new/folder")
```

::: incremental
1. Create a new empty folder.
2. Open RStudio outside any active project.
3. Run `esqlabsR::initProject(destination = "path/to/your/new/folder")` — files and folders are created.
4. In RStudio: `File > New Project > Existing Directory` and select the new folder.
:::

# `ProjectConfiguration`

## Create a `ProjectConfiguration`

The `ProjectConfiguration` object is the R representation of `ProjectConfiguration.xlsx`.

Build one with `createProjectConfiguration()`:

```r
myProjectConfig <- createProjectConfiguration(
  path = "path/to/ProjectConfiguration.xlsx"
)
```

## Modify a `ProjectConfiguration`

Two ways to change a configuration:

- Edit `ProjectConfiguration.xlsx` **before** creating the R object.
- Mutate the R object **after** creation via active fields.

```r
myProjectConfig$outputFolder <- "anotherOutputFolder"
print(myProjectConfig)
```

```{.r code-line-numbers="11"}
<ProjectConfiguration>
  • Working Directory: ~
  • Project Configuration file stored at: ~/testProject/ProjectConfiguration.xlsx

── Paths ──

Folders:
  • Configurations Folder: ~/testProject/Configurations
  • Model Folder: ~/testProject/Models/Simulations
  • Data Folder: ~/testProject/Data
  • Output Folder: ~/testProject/anotherOutputFolder
  • Populations Folder: ~/testProject/Configurations/PopulationsCSV
Files:
  • Model Parameters File: ~/testProject/Configurations/ModelParameters.xlsx
  • Individuals File: ~/testProject/Configurations/Individuals.xlsx
  • Populations File: ~/testProject/Configurations/Populations.xlsx
  • Scenarios File: ~/testProject/Configurations/Scenarios.xlsx
  • Applications File: ~/testProject/Configurations/Applications.xlsx
  • Plots File: ~/testProject/Configurations/Plots.xlsx
  • Data File: ~/testProject/Data/TestProject_TimeValuesData.xlsx
  • Data Importer Configuration File: ~/testProject/Data/esqlabs_dataImporter_configuration.xml
```

# Exercises

## Exercise Overview

1. Create a folder `esqLabsR_training` and initialize a project there.
2. Inspect the resulting layout in your file browser and in RStudio.
3. Switch the project to use a custom `Configurations_V2/` folder — first via the Excel file, then via the R object.

## Initialize the Project

::: {.callout-tip collapse="true"}
## Solution

1. Create a new folder called `esqLabsR_training`.
2. Open RStudio outside any active project.
3. Run:

```r
esqlabsR::initProject(destination = "path/to/esqLabsR_training")
```

4. `File > New Project > Existing Directory` and select `esqLabsR_training`.
:::

## Switch Configurations Folder

We want to point the project at a renamed `Configurations_V2/` folder.

1. Create the new folder `Configurations_V2`.
2. Update `ProjectConfiguration.xlsx` so it points at `Configurations_V2`.
3. Build the `ProjectConfiguration` R object.

```{r, echo = TRUE, eval = FALSE}
#| code-fold: true
#| code-summary: "Solution"
myProjectConfig <- createProjectConfiguration(
  path = "myNewProject/ProjectConfiguration.xlsx"
)

myProjectConfig
# `Configurations_V2` is now linked
```

. . .

4. Change your mind: rewire the folder directly on the existing `ProjectConfiguration` object.

::: {.callout-tip collapse="true"}
## Hint

Active fields are accessed with `$`. The relevant field is `configurationsFolder`.
:::

```{r, echo = TRUE, eval = FALSE}
#| code-fold: true
#| code-summary: "Solution"
myProjectConfig$configurationsFolder <- "Configurations_V2"
```

# Next Steps

## Continue with B3

Scenarios are configured in the Excel files inside `Configurations/`. Continue with the next deck:

- [B3 — Scenario Design](B3-esqlabsR-scenario-design.qmd)

## Resources

- [`{esqlabsR}` documentation website](https://esqlabs.github.io/esqlabsR/)
- [Project structure article](https://esqlabs.github.io/esqlabsR/articles/project-structure.html)
- [`{esqlabsR}` code repository](https://github.com/esqLABS/esqlabsR/)
