---
title: "Sensitivity Analysis"
subtitle: "Introduction to `{esqlabsR}`"
module: "Introduction to `{esqlabsR}`"
ID: B5
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format: esqlabs-revealjs
---

::: r-fit-text
## 🎯 Goal: Quantify parameter influence on PK outcomes {.center}
:::

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.showtext = TRUE,
  fig.align = "center",
  out.width = "70%",
  warning = FALSE,
  message = FALSE
)
library(esqlabsR)
```

# Introduction

## Why sensitivity analysis?

- Quantify how a change in an input parameter shifts a PK outcome (C_max, t_max, AUC, ...).
- Rank parameters by influence — focus model refinement on what matters.
- Detect non-linear behaviour (e.g. saturation).

## Workflow

```{mermaid}
flowchart LR
  A["readScenarioConfigurationFromExcel()"] --> B["createScenarios()"]
  B --> C["scenarios$X$simulation"]
  C --> D["sensitivityCalculation()"]
  D --> E1["sensitivitySpiderPlot()"]
  D --> E2["sensitivityTornadoPlot()"]
  D --> E3["sensitivityTimeProfiles()"]
```

::: {.callout-note title="`{esqlabsR}` vs `{ospsuite}` sensitivity analysis"}
`{ospsuite}::sensitivityAnalysis()` performs a **local** sensitivity analysis: each parameter perturbed once around its nominal value, single sensitivity index per `(parameter, PK metric)` pair.

`{esqlabsR}::sensitivityCalculation()` extends this with:

- multi-factor scaling sweep (`variationRange`) → reveals non-linear / saturation behaviour,
- custom output functions beyond `C_max` / `t_max` / `AUC_inf`,
- ready-made spider / tornado / time-profile plots,
- `saveSensitivityCalculation()` / `loadSensitivityCalculation()` for expensive runs.
:::

## Sensitivity index

For a parameter `p` and PK outcome `y`:

$$
S = \frac{\Delta y / y}{\Delta p / p}
$$

- `S = 1` → 1 % parameter change ⇒ 1 % PK change.
- `S = 0` → no impact.
- Sign → direction (positive: same direction; negative: inverse).

# Setup

## Build a scenario

`sensitivityCalculation()` operates on a `Simulation` object. In the `{esqlabsR}` workflow the simulation is created by a **scenario** — reuse the same path as B3 / B4 instead of loading the `.pkml` directly.

```{r build-scenario}
#| code-fold: show
projectConfigurationPath <- exampleProjectConfigurationPath()

projectConfiguration <- createProjectConfiguration(
  projectConfigurationPath,
  ignoreVersionCheck = TRUE
)

scenarioConfigurations <- readScenarioConfigurationFromExcel(
  scenarioNames = "TestScenario",
  projectConfiguration = projectConfiguration
)

scenarios <- createScenarios(scenarioConfigurations)
```

## Grab the underlying simulation

Each `Scenario` exposes its loaded simulation (with parameters, individual, protocol, and time intervals already applied) via `$simulation`.

```{r get-sim}
#| code-fold: show
simulation <- scenarios$TestScenario$simulation
```

::: {.callout-tip}
List every parameter path of the loaded simulation with `ospsuite::getAllParameterPathsIn(simulation)`.
:::

# Define inputs

## Output paths

Simulation output paths for which sensitivity will be assessed. Each PK parameter (or custom output function) is calculated **for every output path** — one sensitivity result set per `(output path, PK parameter)` combination.

Named vector — names appear as labels in plots.

```{r outputs}
outputPaths <- c(
  Plasma = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
)
```

::: {.callout-tip}
Navigate the simulation tree with `ospsuite::getSimulationTree(simulation)` to find output paths.
:::

## Parameter paths

Example - vary four PBPK-relevant inputs.

```{r params}
parameterPaths <- c(
  Lipophilicity = "Aciclovir|Lipophilicity",
  `Fraction unbound` = "Aciclovir|Fraction unbound (plasma, reference value)",
  Dose = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  `Tubular secretion` = "Neighborhoods|Kidney_pls_Kidney_ur|Aciclovir|Renal Clearances-TS-Aciclovir|Tubular secretion"
)
```

::: {.callout-tip}
Names of the vector → labels on plots / tables. Pick descriptive ones.
:::

::: {.callout-tip}
Find alle parameters that could be used in sensitivity analysis with `ospsuite::getAllParameterPathsIn(simulation)`.
:::

# Run sensitivity

## `sensitivityCalculation()`

```r
analysis <- sensitivityCalculation(
  simulation     = simulation,
  outputPaths    = outputPaths,
  parameterPaths = parameterPaths,
  variationRange = c(0.1, 0.5, 1, 2, 10),
  pkParameters   = c("C_max", "t_max", "AUC_inf")
)
```

What it does:

- Re-runs the simulation once per `(parameter, factor)` pair.
- Scales each parameter by every value in `variationRange`.
- Computes the requested PK metrics + the sensitivity index.

## Run it

```{r run-sens}
#| code-fold: show
analysis <- sensitivityCalculation(
  simulation = simulation,
  outputPaths = outputPaths,
  parameterPaths = parameterPaths
)
```

## Inspect `analysis$pkData`

```{r inspect}
head(analysis$pkData)
```

Columns: `ParameterPath`, `ParameterFactor`, `ParameterValue`, `PKParameter`, `PKParameterValue`, `PKPercentChange`, `SensitivityPKParameter`.

# Visualise

## Spider plot

PK metric on y, scaling factor on x — one panel per metric, one line per parameter.

```{r spider}
sensitivitySpiderPlot(analysis)
```

## Tornado plot

Rank parameters by influence at a fixed scaling — bars compare `0.1×` vs `10×` by default.

```{r tornado}
sensitivityTornadoPlot(analysis)
```

## Time profiles

Full concentration–time curve per scaling factor (blue → black → red as factor increases).

```{r time-profiles}
sensitivityTimeProfiles(analysis)
```

# Beyond defaults

## Custom output functions

`customOutputFunctions` — named list of functions with parameters `x` (time), `y` (output values), or both. Each must return a single number.

```{r custom-fns-def}
#| code-fold: show
customOutputFunctions <- list(
  MeanConcentration = function(y) mean(y),
  NormalizedConcentration = function(y) mean(y / max(y))
)
```

::: {.callout-tip}
- `x` — simulation time vector (minutes); `y` — output vector (base unit of the output)
- Pass `pkParameters = character(0)` to compute **only** custom metrics.
:::

## Run with custom functions

```{r custom-analysis}
#| code-fold: show
analysis_custom <- sensitivityCalculation(
  simulation = simulation,
  outputPaths = outputPaths,
  parameterPaths = parameterPaths,
  pkParameters = character(0),
  customOutputFunctions = customOutputFunctions
)

head(analysis_custom$pkData)
```

Custom metrics appear in the `PKParameter` column — the same structure as standard PK parameters.

## Visualise custom results

```{r custom-tornado}
sensitivityTornadoPlot(analysis_custom)
```

## Save / load

Sensitivity runs are expensive — persist them.

```r
saveSensitivityCalculation(analysis, outputDir = "sensitivity-results")

# Later session
analysis <- loadSensitivityCalculation(
  outputDir  = "sensitivity-results",
  simulation = simulation
)
```

# Exercises

## Exercise 1: Build the scenario

1. Set `projectConfigurationPath` to `exampleProjectConfigurationPath()`.
2. Build `projectConfiguration` with `createProjectConfiguration(..., ignoreVersionCheck = TRUE)`.
3. Read the `TestScenario` configuration with `readScenarioConfigurationFromExcel()`.
4. Realise it via `createScenarios()`.
5. Pull `simulation <- scenarios$TestScenario$simulation` and print it.

```{r ex1}
#| code-fold: true
#| code-summary: "Solution"
projectConfigurationPath <- exampleProjectConfigurationPath()

projectConfiguration <- createProjectConfiguration(
  projectConfigurationPath,
  ignoreVersionCheck = TRUE
)

scenarioConfigurations <- readScenarioConfigurationFromExcel(
  scenarioNames = "TestScenario",
  projectConfiguration = projectConfiguration
)

scenarios <- createScenarios(scenarioConfigurations)

simulation <- scenarios$TestScenario$simulation

simulation
```

## Exercise 2: Run a sensitivity calculation

Vary lipophilicity, fraction unbound, dose, tubular secretion — track plasma.

1. Build the named `outputPaths` and `parameterPaths` vectors.
2. Call `sensitivityCalculation()` with defaults — assign to `analysis`.
3. `head(analysis$pkData)`.

```{r ex2}
#| code-fold: true
#| code-summary: "Solution"
outputPaths <- c(
  Plasma = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
)

parameterPaths <- c(
  Lipophilicity = "Aciclovir|Lipophilicity",
  `Fraction unbound` = "Aciclovir|Fraction unbound (plasma, reference value)",
  Dose = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  `Tubular secretion` = "Neighborhoods|Kidney_pls_Kidney_ur|Aciclovir|Renal Clearances-TS-Aciclovir|Tubular secretion"
)

analysis <- sensitivityCalculation(
  simulation = simulation,
  outputPaths = outputPaths,
  parameterPaths = parameterPaths
)

head(analysis$pkData)
```

::: {.callout-tip}
Path mismatch errors → confirm with `ospsuite::getAllParameterPathsIn(simulation)` and `ospsuite::getAllStateVariablesPaths(simulation)`.
:::

## Exercise 3: Spider plot

1. Generate a spider plot of `analysis`.
2. Identify the PK metric most affected by `Tubular secretion`.

```{r ex3}
#| code-fold: true
#| code-summary: "Solution"
sensitivitySpiderPlot(analysis)
```

## Exercise 4: Tornado plot

1. Generate the tornado plot.
2. Rank parameters by their influence on `AUC_inf`.

```{r ex4}
#| code-fold: true
#| code-summary: "Solution"
sensitivityTornadoPlot(analysis)
```

## Exercise 5: Time profiles + interpretation

1. Plot the time profiles.
2. Which parameter has the strongest visual effect on the terminal slope of the plasma curve? Why?

```{r ex5}
#| code-fold: true
#| code-summary: "Solution"
sensitivityTimeProfiles(analysis)
```

::: {.callout-tip}
Terminal slope reflects elimination — parameters touching clearance pathways (e.g. tubular secretion) dominate it.
:::

## Exercise 6 (optional): Persist results

1. Save `analysis` to `sensitivity-results/` with `saveSensitivityCalculation()`.
2. In a new R session: rebuild the scenario (`createScenarios()`), pull `simulation` from it, then reload with `loadSensitivityCalculation()`.
3. Verify by re-running `sensitivitySpiderPlot()` on the loaded object.

```{r ex6}
#| code-fold: true
#| code-summary: "Solution"
saveSensitivityCalculation(
  analysis,
  outputDir = "sensitivity-results",
  overwrite = TRUE
)

# Fresh session — rebuild the scenario, then reload the analysis.
scenarios <- createScenarios(scenarioConfigurations)
simulation <- scenarios$TestScenario$simulation

analysis <- loadSensitivityCalculation(
  outputDir = "sensitivity-results",
  simulation = simulation
)

sensitivitySpiderPlot(analysis)
```

## Exercise 7 (optional): Custom output function

1. Define a custom function `MaxEliminationRate` returning the steepest (most negative) slope of the concentration–time curve.
2. Run `sensitivityCalculation()` with `pkParameters = character(0)` and only that function.
3. Generate a tornado plot — does the ranking match Exercise 4?

```{r ex7}
#| code-fold: true
#| code-summary: "Solution"
customFunctions7 <- list(
  MaxEliminationRate = function(x, y) {
    slopes <- diff(y) / diff(x)
    abs(min(slopes))
  }
)

analysis7 <- sensitivityCalculation(
  simulation = simulation,
  outputPaths = outputPaths,
  parameterPaths = parameterPaths,
  pkParameters = character(0),
  customOutputFunctions = customFunctions7
)

sensitivityTornadoPlot(analysis7)
```

## Resources

- [`{esqlabsR}` sensitivity article](https://esqlabs.github.io/esqlabsR/articles/sensitivity.html)
- [`{ospsuite}` PK parameters reference](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/pkParameters.html)
- [`{esqlabsR}` code repository](https://github.com/esqLABS/esqlabsR/)
