---
title: "Quarto Reporting for M&S Projects"
subtitle: "Reproducible reports from `{esqlabsR}` projects"
module: "Quarto Reporting"
ID: D1
date: last-modified
institute: |
  ![](../resources/esqlabs.svg){height=100}
format:
  esqlabs-revealjs:
    scrollable: true
    code-overflow: scroll
---

::: {.r-fit-text}
## 🎯 Goal: Generate a regulatory-grade report end-to-end from your project files
:::

# Why Quarto

## The reproducibility gap

::: {.incremental}
- Word docs: figures pasted, tables hand-typed → drift from results
- Slides: same problem, plus auto-formatting tricks
- A regulator asks: "show the same plot for 15 mg" → days of work
- Quarto: the report **is** code. Re-render → fresh figures + tables.
:::

## What Quarto is

::: {.incremental}
- Successor to R Markdown — language-agnostic
- One source → HTML, PDF, Word, reveal.js, dashboard, …
- Run R / Python / Julia / Observable in chunks
- Parameters, citations, cross-refs, equations all built-in
:::

## Why pair it with `{esqlabsR}`

::: {.incremental}
- The project folder already has scenarios, plots, SA results — defined in Excel, executed in R
- Quarto stitches them into a narrative
- Re-render = re-run scenarios + re-render plots + regenerate tables
- Output: HTML for review, PDF for submission, both from the same source
:::

# Quarto basics

## Anatomy of a `.qmd` file

```yaml
---
title: "Midazolam PBPK — Single-Dose"
author: "Pavel Balazki"
date: today
format:
  html:
    toc: true
  pdf:
    documentclass: scrreprt
---
```

```` markdown
# Background

Some markdown text. References work via `@author2024`.

```{r setup}
#| include: false
library(esqlabsR)
project <- createProjectConfiguration("ProjectConfiguration.xlsx")
```

## Results

```{r}
#| label: fig-time-profile
#| fig-cap: "Plasma concentration vs observed."
scenarioConfig <- readScenarioConfigurationFromExcel(
  "Midazolam_PO_7.5mg",
  projectConfiguration = project
)
scenarios <- createScenarios(scenarioConfig)
results   <- runScenarios(scenarios = scenarios)
plots     <- createPlotsFromExcel(plotGridNames        = "Midazolam_PO_vs_obs",
                                  simulatedScenarios   = results,
                                  projectConfiguration = project)
plots[[1]]
```
````

## Code chunk options that matter

| Option | Effect |
|---|---|
| `#| echo: false` | hide code |
| `#| include: false` | run, hide both code + output |
| `#| eval: false` | show code, don't run |
| `#| cache: true` | cache result; re-use on next render |
| `#| fig-cap: "..."` | figure caption |
| `#| label: fig-foo` | cross-ref via `@fig-foo` |
| `#| tbl-cap: "..."` | table caption |

## Parameters

```yaml
---
params:
  dose_mg: 7.5
  route: "PO"
  scenario: "Midazolam_PO_7.5mg"
---
```

```{.r}
# inside a chunk
params$dose_mg
```

Re-render with new params:

```bash
quarto render Reports/Midazolam-report.qmd -P dose_mg:15
```

. . .

::: {.center-x}
**The same `.qmd` becomes a 7.5 mg report and a 15 mg report.**
:::

## Rendering

```bash
# from the project root
quarto render Reports/Midazolam-report.qmd                 # default format(s)
quarto render Reports/Midazolam-report.qmd --to html       # HTML only
quarto render Reports/Midazolam-report.qmd --to pdf        # PDF only
quarto render Reports/Midazolam-report.qmd -P dose_mg:15   # parameter override
```

. . .

::: {.callout-tip}
PDF rendering needs LaTeX. Use TinyTeX: `quarto install tinytex`.
:::

# A regulatory-style report — section by section

## Recommended structure

1. **Background & objectives**
2. **Model description** (compound, structure, references)
3. **Data sources** (clinical / preclinical observations)
4. **Scenario definitions** (auto-tabulated from `Scenarios.xlsx`)
5. **Simulation results** (figures via `createPlotsFromExcel`)
6. **PK parameter analysis** (table via `calculatePKAnalyses`)
7. **Sensitivity analysis** (load from `Results/SA_*.rds`)
8. **Discussion + Conclusions**
9. **Appendix:** session info, package versions, hash of the model file

## 1. Background — pure markdown

```markdown
# Background

Midazolam is a CYP3A4 substrate widely used as a probe for studying
clearance pathways and drug-drug interactions [@gorski1998].

## Objectives

This report presents single-dose PBPK simulations of midazolam after
oral administration of `{r} params$dose_mg` mg, compared against the
observations of @greenblatt1989.
```

::: {.fragment}
**Inline R via `` `{r} ...` ``** — the dose is read from `params`. No copy-paste.
:::

## 4. Auto-tabulating scenario specs

```{.r}
#| label: tbl-scenarios
#| tbl-cap: "Scenario specifications."
library(readxl); library(knitr)

read_excel(file.path(project$paths$ConfigurationsFolder, "Scenarios.xlsx"),
           sheet = "Scenarios") |>
  dplyr::filter(Scenario == params$scenario) |>
  knitr::kable()
```

## 5. Figure grid

```{.r}
#| label: fig-results
#| fig-cap: "Simulation vs observed."
plots <- createPlotsFromExcel(
  plotGridNames        = "Midazolam_PO_vs_obs",
  simulatedScenarios   = results,
  observedData         = observedData,
  projectConfiguration = project
)
plots$Midazolam_PO_vs_obs
```

## 6. PK parameter table

```{.r}
#| label: tbl-pk
#| tbl-cap: "Calculated PK parameters."
pk <- calculatePKAnalyses(results = results[[1]])
pkDF <- pkAnalysesToDataFrame(pk)
# optionally filter by QuantityPath:
# pkDF <- pkDF[pkDF$QuantityPath == "Organism|...|Midazolam|Concentration", ]

pkAnalysesAsDataFrame(pk) |>
  dplyr::select(Parameter, Value, Unit) |>
  knitr::kable(digits = 3)
```

## 7. Sensitivity from saved results

```{.r}
#| label: fig-sa
#| fig-cap: "Tornado plot of CYP3A4 kcat, lipophilicity, GFR-fraction."
analysis <- loadSensitivityCalculation(
  outputDir  = file.path(project$paths$ResultsFolder, "SA_Midazolam"),
  simulation = scenarios[[1]]$simulation
)
sensitivityTornadoPlot(analysis)
```

## 9. Appendix — make the report auditable

```{.r}
#| label: appendix
#| echo: false
sessionInfo()                                  # R + package versions
tools::md5sum("Models/Simulations/Midazolam_PO_7.5mg.pkml")  # hash the model file
Sys.time()                                     # render time
```

::: {.center-x}
A regulator can verify: same model file, same package versions → same numbers.
:::

# Best practices

## DO

::: {.incremental}
- Cache slow chunks (`#| cache: true`) — but invalidate when inputs change
- Use `#| label:` on every figure / table — enables cross-refs
- Render PDF + HTML together — HTML for review, PDF for archive
- Pin package versions via `renv::snapshot()`
- Commit the `.qmd` (not the rendered output)
:::

## DON'T

::: {.incremental}
- Don't paste figures from somewhere else — always regenerate
- Don't hard-code paths — use `here::here()` or `project$paths$...`
- Don't skip `set.seed(...)` if any chunk uses randomness
- Don't write hundreds of lines of R inside chunks — refactor into the project's `R/` folder, then `source()` them
- Don't render with uncommitted changes if you ship the PDF (irreproducible)
:::

# 🖐️ Hands-on

## Tasks (10 min)

1. Open `Reports/Midazolam-report.qmd` in RStudio
2. Click **Render** → inspect the HTML
3. From the terminal, render with a different dose:
   ```
   quarto render Reports/Midazolam-report.qmd -P dose_mg:15
   ```
4. **Bonus:** add a new chunk that prints `Sys.time()` in the appendix and re-render
5. **Bonus:** change the title via `params$title` (add it to the YAML)

::: {.center-x}
⏰ **10 min** — solutions in `Reports/Midazolam-report-solution.qmd`
:::

# Wrap-up

## What you have now

::: {.incremental}
- A reproducible report driven by a parameterized `.qmd`
- Figures + tables that regenerate from raw project files
- HTML for review + PDF for archive — same source
- An auditable appendix with package versions + model hash
:::

. . .

::: {.center-x .r-fit-text}
**Your project folder + one `.qmd` = a re-runnable submission package.**
:::

## Further reading

- Quarto docs — `quarto.org`
- `{esqlabsR}` reporting cookbook — `esqlabs.com/r-trainings`
- Parameterized reports — `quarto.org/docs/computations/parameters.html`
- `{officedown}` for Word output styling
- `{targets}` if your pipeline grows complex
