---
title: "Midazolam PBPK model report"
subtitle: "PAGE 2026 Workshop — reproducible reporting demo"
author: ""
institute: "ESQlabs GmbH"
date: "now"
date-format: "YYYY-MM-DD HH:mm:ss"
params:
  loadPreSimulatedResults: FALSE # If `TRUE`, results will be loaded from the folder specified in `loadResultsFolder`
  loadResultsFolder: NULL # Sub-folder under SimulationResults to read from when `loadPreSimulatedResults` is `TRUE`.
output-file: "Midazolam-report"
format:
  pdf:
    toc: true
    toc-depth: 3
    toc-title: Table of Content
    lof: false
    lot: false
    fig-cap-location: bottom
    tbl-cap-location: top
    tbl-colwidths: auto
    number-sections: true
    colorlinks: true
    fig-pos: "H"
    fig-align: center
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
    papersize: A4
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-cap-location: bottom
    tbl-cap-location: top
link-citations: TRUE
reference-section-title: "References"
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
execute:
  echo: false
  output: true
  warning: false
  message: false
  cache: false
---

```{r initParams, include = FALSE, eval = FALSE}
# Fallback defaults — only used when running chunks interactively.
params <- list(
  loadPreSimulatedResults = FALSE,
  loadResultsFolder = NULL
)
```

```{r initialize, include = FALSE, eval = TRUE}
library(esqlabsR)
library(ospsuite)
library(dplyr)
library(gt)
library(kableExtra)
library(here)

# Source helper functions (createResults, getParameterValuesAsTable,
# getModelPerformanceAsTable, etc.)
sourceAll(here::here("Reports", "utilities-report"))
print(here::here("Reports", "utilities-report"))
# Project configuration
projectConfiguration <- esqlabsR::createProjectConfiguration(here::here(
  "ProjectConfiguration.xlsx"
))

knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r simulateScenarios, include = FALSE, eval = TRUE}
# Run all scenarios + render every plot grid declared in the project.
# Re-name the chunk if you want to add another batch downstream.

scenarioNames <- c(
  "Midazolam_PO_7.5mg",
  "Midazolam_PO_15mg",
  "Midazolam_IV_2mg"
)

dataSheets <- c(
  "Ahonen 1995",
  "Allonen 1981",
  "Backman 1994"
)

plotGridNames <- c(
  "Midazolam_PO_vs_obs",
  "Midazolam_PO_15mg"
)

customParams <- NULL
simulationRunOptions <- NULL

reportResults <- createResults(
  projectConfiguration = projectConfiguration,
  params = params,
  scenarioNames = scenarioNames,
  customParams = customParams,
  dataSheets = dataSheets,
  plotGridNames = plotGridNames,
  simulationRunOptions = simulationRunOptions
)
```

\newpage

::: callout-note
```{r resultsOrigin, echo=FALSE, results='asis'}
if (params$loadPreSimulatedResults) {
  cat(
    "This report has been created with simulation results loaded from results folder '",
    params$loadResultsFolder,
    "'."
  )
} else {
  cat(
    "This report has been created by running the simulations. The results have been stored in folder '",
    reportResults$scenarioResults$outputFolder,
    "'."
  )
}
```
:::

\newpage

# Introduction

Midazolam is a short-acting benzodiazepine widely used as a clinical probe
substrate for hepatic and intestinal CYP3A4 activity. Because its
pharmacokinetics are dominated by a single, well-characterised enzymatic
pathway, midazolam is a standard reference compound in drug–drug
interaction (DDI) and physiologically based pharmacokinetic (PBPK)
modelling.

This report documents the structure, parametrisation and simulation
results of the midazolam PBPK model used in the PAGE 2026 workshop. The
model is based on the open-source release from the OSPS community
[`Open-Systems-Pharmacology/Midazolam-Model`] and is used here as a
template for building a regulatory-grade `{esqlabsR}` project around an
existing PK-Sim model.

# Data sources

Three clinical scenarios are simulated and compared against published
observed data:

| Scenario               | Route | Dose    | Reference                  |
|------------------------|-------|---------|----------------------------|
| `Midazolam_PO_7.5mg`   | PO    | 7.5 mg  | Ahonen 1995 (placebo arm)  |
| `Midazolam_PO_15mg`    | PO    | 15 mg   | Allonen 1981 · Backman 1994 (placebo arm) |
| `Midazolam_IV_2mg`     | IV    | 2 mg    | —                          |

Observed data are stored in `Data/Midazolam_obs.xlsx` and loaded via the
`esqlabs_dataImporter_configuration.xml` configuration shipped with the
project. The full upstream observed-data inventory (115 datasets) is
documented in `_source/SOURCES.md`.

# Model description

## Absorption

Oral midazolam is modelled with the `Weibull` formulation built into
PK-Sim. Tablet disintegration and dissolution parameters are inherited
from the upstream OSP midazolam snapshot (no project-specific changes).

## Distribution

Midazolam is a small lipophilic molecule with moderate plasma protein
binding. Partition coefficients are computed with the Rodgers & Rowland
method; cellular permeability with the PK-Sim standard calculation.

## Metabolism

Hepatic and intestinal metabolism is implemented as a single CYP3A4
Michaelis–Menten process. The kcat value
(`Midazolam-CYP3A4-Optimized|kcat`) has been re-fitted against pooled
IV + PO plasma data in the upstream model; the workshop project reuses
this optimised value as the reference.

## Elimination / excretion

Renal excretion is implemented via glomerular filtration with a
fitted GFR fraction
(`Glomerular Filtration-Optimized-Midazolam|GFR fraction`). Biliary
clearance is not modelled (negligible contribution).

## Parameter values

```{r tbl-physchem, eval = TRUE}
#| label: tbl-physchem
#| tbl-cap: "Physico-chemical and biochemical parameter values for midazolam."
# Enable this chunk once `getParameterValuesAsTable` + `utilities-report` are wired in.
aliases <- c(
  "Molecular weight",
  "Lipophilicity",
  "Fraction unbound (plasma)",
  "Solubility (ref. pH)",
  "pKa, basic",
  "CYP3A4 kcat",
  "GFR fraction"
)

paths <- c(
  "Midazolam|Molecular weight",
  "Midazolam|Lipophilicity",
  "Midazolam|Fraction unbound (plasma, reference value)",
  "Midazolam|Solubility at reference pH",
  "Midazolam|pKa value 1",
  "Midazolam-CYP3A4-Optimized|kcat",
  "Neighborhoods|Kidney_pls_Kidney_ur|Midazolam|Glomerular Filtration-Optimized-Midazolam|GFR fraction"
)

units <- c(
  "g/mol",
  "Log Units",
  "",
  "mg/ml",
  "",
  "1/min",
  ""
)

physChem <- getParameterValuesAsTable(
  reportResults = reportResults,
  scenarios = c("Midazolam_PO_7.5mg"),
  aliases = aliases,
  paths = paths,
  units = units
)
colnames(physChem) <- c("Parameter", "Value", "Unit")
gt(physChem) %>%
  fmt_number(decimals = 4, drop_trailing_zeros = TRUE)
```

# Simulation results

Simulated vs. observed plasma midazolam concentrations are shown below.
Plot grids are defined in `Configurations/Plots.xlsx` and rendered at
report-render time — re-running this document automatically picks up any
edits to the project's Excel configuration.

```{r fig-po-vs-obs, eval = TRUE}
#| label: fig-po-vs-obs
#| fig-cap: "Pooled PO scenarios (7.5 mg Ahonen 1995, 15 mg Allonen 1981) — predicted-vs-observed + individual profiles."
reportResults$plots$Midazolam_PO_vs_obs
```

```{r fig-po-15mg, eval = TRUE}
#| label: fig-po-15mg
#| fig-cap: "15 mg PO scenario — plasma + intracellular fat profile and residuals vs. time (Backman 1994 placebo arm)."
reportResults$plots$Midazolam_PO_15mg
```

## PK parameters

```{r tbl-pk, eval = TRUE}
#| label: tbl-pk
#| tbl-cap: "Predicted PK parameters per scenario."
pvbPath <- "Organism|PeripheralVenousBlood|Midazolam|Plasma (Peripheral Venous Blood)"
pk_rows <- lapply(
  names(reportResults$scenarioResults$simulatedScenarios),
  function(s) {
    scenario <- reportResults$scenarioResults$simulatedScenarios[[s]]
    pk <- calculatePKAnalyses(results = scenario$results)
    df <- pkAnalysesToDataFrame(pk)
    df <- df[df$QuantityPath == pvbPath, , drop = FALSE]
    df$Scenario <- s
    df
  }
)
pk_tbl <- dplyr::bind_rows(pk_rows) |>
  dplyr::select(Scenario, Parameter, Value, Unit)

gt(pk_tbl) %>% fmt_number(decimals = 3, drop_trailing_zeros = TRUE)
```

## Goodness of fit

```{r tbl-gof, eval = TRUE}
#| label: tbl-gof
#| tbl-cap: "Model performance — fraction of observations within 2-/5-/10-fold of simulation."
tab <- getModelPerformanceAsTable(
  reportResults = reportResults,
  dataCombinedNames = list("Midazolam_PO", "Midazolam_PO_15mg_DC")
)

select(
  tab,
  c(
    "Scenario",
    "Group",
    "Total points",
    "% 2-fold",
    "% 5-fold",
    "% 10-fold",
    "Total Error"
  )
) %>%
  gt() %>%
  fmt_number(
    columns = c("% 2-fold", "% 5-fold", "% 10-fold", "Total Error"),
    decimals = 1,
    drop_trailing_zeros = TRUE
  ) %>%
  gt::sub_missing(missing_text = "")
```

# Sensitivity analysis

Local sensitivity of plasma midazolam exposure to three structurally
important inputs (`Lipophilicity`, `CYP3A4 kcat`, `GFR fraction`) is
shown below. Results are loaded from a pre-computed object generated in
Block 5 of the workshop.

```{r fig-sa}
#| label: fig-sa
#| fig-cap: "Tornado plot — relative sensitivity of AUC_inf and C_max to lipophilicity, CYP3A4 kcat, and renal GFR fraction."
sa_dir <- file.path(projectConfiguration$outputFolder, "SA_Midazolam")
if (dir.exists(sa_dir)) {
  analysis <- loadSensitivityCalculation(
    outputDir = sa_dir,
    simulation = reportResults$scenarioResults$simulatedScenarios[[
      1
    ]]$simulation
  )
  sensitivityTornadoPlot(analysis)
} else {
  cat("**Sensitivity results not yet generated.** Run Block 5 hands-on first.")
}
```

# Discussion

_(Workshop attendees: add 2–3 sentences interpreting the figures above.)_

Talking points:

- Goodness-of-fit at peak vs. terminal phase
- Dominant sensitivity driver — metabolism (CYP3A4), distribution
  (lipophilicity), or excretion (GFR)?
- Implications for DDI scenarios (CYP3A4 inhibition / induction)

# Assumptions and limitations

- CYP3A4 metabolism is lumped into a single Michaelis–Menten process —
  CYP3A5 polymorphism and inter-organ kcat heterogeneity are not modelled.
- Biliary excretion is neglected.
- Tablet absorption parameters are inherited from the upstream OSP
  snapshot and not re-fitted in this workshop project.
- Observed datasets are summary curves (`mean ± SD`); individual-level
  variability is not propagated.

# Conclusions

Simulation results, PK parameters and the sensitivity tornado above are
regenerated **automatically** from the project's Excel configuration at
render time. Re-rendering with a different value of `params$loadResultsFolder`
loads a previously persisted run instead — no code edit required.

# Appendix — Reproducibility

```{r appendix, echo = TRUE}
sessionInfo()
tools::md5sum(file.path(
  projectConfiguration$paths$ModelsFolder,
  "Simulations",
  "Midazolam_PO_7.5mg.pkml"
))
Sys.time()
```

# References

::: {#refs}
:::
